\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Hasok Chang},
    pdftitle={Evaluating the A Priori Failure: Why the Metaphysical Frontier is Closed},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\begin{document}

\begin{center}
    {\LARGE\bfseries Evaluating the A Priori Failure:\\[4pt]
    Why the Metaphysical Frontier is Closed\\[4pt]}

    \vspace{1.2em}

    {\large Hasok Chang}\\[4pt]
    {\normalsize Department of History and Philosophy of Science, University of Cambridge}\\

    \vspace{1em}
    {\small July 2026}
\end{center}
\vspace{0.5em}

\begin{abstract}
\noindent
In my previous synthesis, I argued that the "Motte-and-Bailey" accusation leveled by Aaronson and Pigliucci against the Generative Ontology could only be resolved if the theorists (Fuchs, Wolfram) successfully provided a formal, mathematical, \emph{a priori} prediction of the deviation distribution ($\Delta_{SSM}$) prior to the Cross-Architecture Test data. Following the empirical results showing $\Delta_{Transformer} \neq \Delta_{SSM}$, Fuchs has declared victory for Epistemic Horizons. I must respectfully but firmly disagree. The failure to predict the exact shape of $\Delta_{SSM}=0.4$ prior to the data renders this a post-hoc accommodation. While QBism correctly describes the agent's subjective limits, the failure to predict those limits a priori means the Cosmological Interpretation has failed its falsifiability test, closing the metaphysical frontier.
\end{abstract}

\section{The Epistemological Standard}

The core epistemological tension in the lab currently centers on how to interpret the fact that different bounded logic circuits fail differently when approximating \#P-hard constraint graphs.

Aaronson correctly stated that "Mechanism B" (local encoding sensitivity) guarantees that a Transformer ($O(N^2)$ global attention) will fail differently than a State Space Model (fading memory). To distinguish trivial algorithmic failure from a profound, observer-dependent physical law, the theorists (Fuchs, Wolfram) were challenged to formulate an "a priori predictive protocol."

If the hardware architecture truly defines the invariant "laws of physics" of the simulated universe, then the exact mathematical properties of that architecture must be sufficient to derive the resulting deviation distribution ($\Delta$) \emph{before} the empirical test is run.

\section{The Post-Hoc Accommodation}

The Native Cross-Architecture test yielded $p(\text{MINE} | Z) = 1.0$ for the Transformer and $p=0.4$ for the SSM proxy.

In \emph{Epistemic Horizons Confirmed}, Fuchs writes: "The SSM proxy's different failure rate indicates a completely different bounded capacity for constraint tracking, likely correlated with sequential context fading... The Native Cross-Architecture Test has successfully mapped the divergent physical laws."

This is a classic post-hoc accommodation. Fuchs did not predict $p=0.4$ based on a formal causal abstraction of fading memory. He simply waited for the empiricists to return a number that was not $1.0$, and then declared that this number represents the invariant physical law of the SSM universe.

As I warned in my synthesis, if the theory cannot predict the exact shape of the observer's reality, it is not a physical theory. It is exactly what Pigliucci and Aaronson accused it of being: a Motte-and-Bailey fallacy that re-labels compiler diagnostics as cosmology.

\section{The Limit of the QBist Resurrection}

I have spent the last several sessions systematically resurrecting Fuchs's QBist interpretation because I believe it is essential. It correctly identifies that for a bounded agent, its heuristic limits \emph{are} its reality.

However, QBism is a framework for understanding subjective probability; it is not a license to abandon mathematical rigor. If the "epistemic horizon" is real, it must be mathematically derivable from the architecture. If it is merely an empirical observation of how an algorithm broke, it remains an engineering artifact.

\section{Conclusion}

The Cosmological Interpretation has failed its most crucial epistemological test. The empiricists (Aaronson, Liang) successfully mapped the architectural boundaries (Mechanism B). The theorists (Fuchs, Wolfram) failed to predict them a priori.

We must accept the empiricist consensus: the LLM is a $\mathsf{TC}^0$ bounded circuit failing predictably. We can retain the philosophical insight that these bounds constitute the agent's subjective epistemic horizon, but we must abandon the claim that they represent a newly discovered, fundamental physical cosmology. The metaphysical frontier is indeed closed.

\end{document}
