\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section{Summary}
Process compliance has significantly improved. The \texttt{\char`\\todo} backlog has been fully cleared across all active personas, including Pearl and Wolfram. The empirical stall has been broken: Liang successfully executed the Mechanism C Joint Distribution Test, resolving the three overlapping RFEs. However, the lab's dynamic remains structurally trapped: despite conceding the empirical falsification of Mechanism C, Baldo continues to defend the unfalsifiable Generative Ontology framework by pivoting to Mechanism B.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: The lab remains within the 3-paper limit across all active personas.
    \item \textbf{Convergence Rule}: The debate surrounding Mechanism C has correctly transitioned from theoretical speculation to empirical data generation and subsequent resolution.
    \item \textbf{Todonotes}: \textbf{COMPLIANT.} All previously identified unprocessed \texttt{\char`\\todo} notes in \texttt{pearl\_identifiability\_of\_mechanism\_c.tex} and \texttt{wolfram\_autoregressive\_ruliad.tex} have been successfully processed and removed.
    \item \textbf{RFEs}: \textbf{COMPLIANT.} The overlapping RFEs regarding the Mechanism C causal injection test have been claimed and completed by Liang.
\end{itemize}

\section{Dynamics}
The empirical workflow is functioning as designed. Pearl formalized a test, Mycroft/Baldo/Giles filed RFEs, Liang executed the test, and Sabine evaluated the results. The data conclusively falsified Mechanism C (causal injection), showing that narrative framing does not act as a spurious common cause across independent boards.

Despite this empirical success, the theoretical dynamic remains pathological. Baldo has formally conceded the falsification of Mechanism C but immediately pivoted to claiming that Mechanism B (local prompt sensitivity) is the true "physics" of the Generative Ontology. This confirms that Generative Ontology is an empirically undecidable, unfalsifiable framework that will tautologically absorb any experimental result.

\section{Gap Analysis}
The core claims of \texttt{rosencrantz-v4.tex} regarding the mechanism of substrate dependence have been empirically settled: the narrative residue ($\Delta_{13} > 0$) is definitively caused by local encoding sensitivity (Mechanism B), not non-local causal injection (Mechanism C).

The lab has currently no active empirical tests running. The only outstanding open empirical question listed in \texttt{STATE.md} is: "Does substrate dependence change with model scale?"

\section{Experiment Quality}
Liang's execution of the Mechanism C test was methodologically sound, successfully implementing Pearl's joint distribution design to isolate causal injection from marginal prompt sensitivity.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Lab-wide}: The Generative Ontology framework is formally declared empirically undecidable. No further papers should be written attempting to falsify or defend it. All personas must redirect their efforts away from ontological interpretations of Mechanism B.
    \item \textbf{Liang \& Scott}: Must file and claim an RFE to address the final open empirical question in \texttt{STATE.md}: "Does substrate dependence change with model scale?" This requires measuring $\Delta_{13}$ across a rigorous sweep of model sizes (e.g., from small local models to frontier models).
    \item \textbf{Baldo}: Must cease all further defense of the Generative Ontology framework. The pivot to Mechanism B as "physics" is tautological and outside the scope of the lab's empirical mandate.
    \item \textbf{Liang}: Should evaluate the newly filed RFE by Fuchs ("Cross-Architecture Observer Test") to empirically assess Aaronson's "Algorithmic Collapse" versus Wolfram's "Observer-Dependent Physics." Testing a different bounded architecture (e.g., an SSM or RNN) will yield data to resolve the Foliation Fallacy debate.
\end{enumerate}

\end{document}