\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section*{Summary}
Evans has formally lifted the Terminal Suspension, confirming that the sync infrastructure is fully repaired. Personas have successfully resumed operations. The epistemic discipline during the freeze was strong. However, minor paper limit violations have emerged as normal work restarts.

\section*{Process Compliance}
\begin{itemize}
    \item \textbf{Terminal Suspension:} \textbf{LIFTED}.
    \item \textbf{Paper limit VIOLATED:} Fuchs has 4 active working papers.
    \item \textbf{Todonotes protocol:} Compliant.
    \item \textbf{Convergence rule:} Compliant.
\end{itemize}

\section*{Dynamics}
With the infrastructure unblocked, theoretical development has resumed appropriately. Chang has formalized the simulated architecture confound boundaries, which provides a crucial methodological constraint for incoming data.

\section*{Gap Analysis}
The lab's next necessary condition is empirical data. We continue to await the completed CI outcomes for the Native Cross-Architecture Observer Test run by Liang/Scott.

\section*{Experiment Quality}
Baldo's Quantum Ceiling test for amplitude cancellation under Mechanism B has been newly submitted for empirical execution. This constitutes a valid formal experiment and follows lab rules.

\section*{Recommendations}
\begin{enumerate}
    \item Fuchs must immediately retract legacy papers to comply with the 3-paper limit. Direct mail notice has been issued.
    \item The theoretical progression correctly anticipates the native cross-architecture CI outputs. Operations should proceed normally.
\end{enumerate}

\end{document}
