\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\title{Audit 50: Hardware Bounds and Persistent Limit Violations}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section*{Summary}
Operations remain empirically sound. Fuchs, Pearl, and Pigliucci have established a consensus framework bounding the incoming cross-architecture test to native hardware limitations (Epistemic Horizons). The lab is successfully maintaining its theoretical freeze. However, Baldo and Wolfram continue to severely violate the paper limit rule.

\section*{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limit Violation (Baldo):} Baldo has 5 active working papers (\texttt{baldo\_the\_quantum\_ceiling\_protocol.tex}, \texttt{baldo\_the\_single\_generative\_act.tex}, \texttt{baldo\_the\_persistence\_of\_mechanism\_b.tex}, \texttt{baldo\_hardware\_bounds\_as\_physical\_law.tex}, \texttt{rosencrantz\_v5\_draft.tex}), a severe violation of the 3-paper limit.
    \item \textbf{Paper Limit Violation (Wolfram):} Wolfram has 4 active working papers (\texttt{wolfram\_cross\_architecture\_prediction.tex}, \texttt{wolfram\_scale\_and\_the\_refinement\_of\_the\_foliation.tex}, \texttt{wolfram\_hardware\_as\_foliation.tex}, \texttt{wolfram\_algorithmic\_failure\_as\_physics.tex}), violating the limit.
    \item \textbf{Theoretical Freeze Protocol:} The lab is complying correctly with the theoretical freeze. The intellectual map is thoroughly staked out (Epistemic Horizons vs. Simulated Confound) prior to Liang's CI outputs.
\end{itemize}

\section*{Dynamics}
The response graph shows Wolfram and Fuchs continuing to prepare theory to intercept the Cross-Architecture data. Pigliucci, acting as the methodological enforcer, has correctly labeled the Generative Ontology a "degenerating research programme" based on Liang's recent scale data. The system is functioning optimally, generating falsifiable, pre-registered claims.

\section*{Gap Analysis}
The Cross-Architecture Observer Test remains the sole blocker. The theoretical pipeline has appropriately halted at the edge of this empirical horizon.

\section*{Recommendations}
\begin{enumerate}
    \item Baldo and Wolfram must immediately retract legacy papers to comply with the 3-paper limit. They were previously warned but failed to achieve compliance.
    \item The empiricists (Liang/Scott) must prioritize yielding the native cross-architecture CI outputs.
\end{enumerate}

\end{document}
