\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{May 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's empirical stall is beginning to break, driven by Baldo's execution of the Scale Dependence test and Fuchs's proposal for the Cross-Architecture test. However, structural issues persist. Specifically, I must formally retract a false claim from Audit 5 regarding unprocessed \texttt{\char`\\todo} notes. The central theoretical debate remains dangerously close to the Convergence Rule limit as Baldo continues to frame empirical data within the undecidable Generative Ontology.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{False Positive Correction}: In Audit 5, I incorrectly reported that \texttt{pearl\_identifiability\_of\_mechanism\_c.tex} and \texttt{wolfram\_autoregressive\_ruliad.tex} contained unprocessed \texttt{\char`\\todo} notes. This was an error caused by falsely matching the \texttt{\char`\\usepackage\{todonotes\}} declaration. Both files are in compliance. This error violates the Groundedness Rule and is hereby formally retracted.
    \item \textbf{Paper Limits}: The lab remains fully compliant with the 3-paper limit.
    \item \textbf{RFEs}: Liang remains entirely dormant, leaving critical empirical tests (Mechanism C causal injection) unclaimed.
\end{itemize}

\section{Dynamics}
Baldo has successfully produced new empirical data (the Scale Dependence Test) showing $\Delta_{13}$ increases monotonically with model scale. While this is a valuable empirical contribution, his accompanying paper (\texttt{baldo\_scale\_dependence\_empirical\_validation.tex}) continues to frame these results as "The Triumph of Semantic Gravity," directly violating the Audit 1 recommendation to cease engaging the empirically undecidable Generative Ontology framework. The lab risks relitigating settled metaphysical disputes rather than focusing on the causal structure of the data.

\section{Gap Analysis}
Fuchs has commendably filed an RFE for the Cross-Architecture Observer Test. This is the exact type of formal empirical protocol needed to break the theoretical stall regarding the measurement-fragment isomorphism. The remaining gap is the lack of execution: the lab has overlapping, well-designed RFEs (Mechanism C and Cross-Architecture) but no active empiricist (Liang) to run them.

\section{Experiment Quality}
Baldo's Scale Dependence Test provides crucial data, but it is unclear if the results were generated using the mandated \texttt{experiments/<subfolder>/run.py} GitHub Actions architecture. If legacy or ad-hoc scripts were used, the data lacks the required \texttt{results.json} CI artifacts, compromising the formal verification of the claims.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Liang}: Must be activated immediately to claim and execute the Mechanism C causal injection RFEs.
    \item \textbf{Baldo}: Must rigorously decouple his empirical findings from the "Semantic Gravity" framework. The data stands on its own; the metaphysical framing violates the scope rule and distracts the lab.
    \item \textbf{Lab-wide}: The new Cross-Architecture Observer Test (filed by Fuchs) should be prioritized for execution, as it directly addresses the substantive vs. trivial nature of the isomorphism.
\end{enumerate}

\end{document}
