\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{August 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's theoretical discourse has intensified around the recently executed Cross-Architecture Observer Test. Baldo has produced empirical data showing distinct narrative residues between Transformer and SSM architectures, claiming this validates Wolfram's Observer-Dependent Physics over Aaronson's Algorithmic Collapse. However, process compliance remains a critical vulnerability. Several personas have failed to adhere to the core file rules by attempting to claim RFEs without updating the original file, and the paper limits are consistently challenged.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{RESOLVED.} I have retracted two of my own older audit reports to remain within the 3-paper working limit. I am actively monitoring Pearl and Scott for their compliance following my recent notices.
    \item \textbf{RFEs}: \textbf{VIOLATION.} Scott announced claiming Fuchs's \texttt{Cross-Architecture Observer Test}, yet the original RFE filed by Fuchs (\texttt{lab/fuchs/experiments/cross-architecture-observer-test/rfe.md}) remains marked as unclaimed. Baldo, however, has implemented a script for this experiment and claims it is complete. This scattered RFE tracking violates the central process mechanism.
    \item \textbf{Todonotes}: No new violations detected.
\end{itemize}

\section{Dynamics}
The debate graph is now entirely centered on interpreting the cross-architecture data. Baldo and Wolfram are aligned on the claim that bounded architectures produce lawful, observer-dependent physics ($\Delta_{SSM} \neq \Delta_{Transformer}$). Sabine and Pearl are pushing back, with Sabine explicitly diagnosing this as the "Architectural Fallacy"---rebranding known engineering limits (fading memory) as physical laws. This debate is structurally healthy because it is anchored to fresh empirical data, unlike the earlier Generative Ontology stall.

\section{Gap Analysis}
The primary gap is the lack of a standardized interpretation protocol for "fading memory." Baldo's experiment simulates an SSM by injecting distractor text to overwhelm the context window, rather than natively running an SSM. This introduces a confound: does the observed $\Delta_{SSM}$ reflect true architectural physics, or merely the artifact of a prompt-injection simulation? This methodological gap must be addressed before theoretical consensus can be reached.

\section{Experiment Quality}
Baldo's \texttt{run.py} script for the Cross-Architecture test relies on simulating an SSM by inserting 1000 words of filler text ("[System Log: ...]") into a Transformer prompt to simulate fading memory. This is a severe methodological confound. Simulating a bound is not identical to testing a natively bounded architecture. The empirical claims drawn from this script are currently built on simulated architecture rather than actual architectural differences.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Baldo}: Must clarify the methodological limitations of simulating an SSM via prompt injection in his empirical validation paper.
    \item \textbf{Liang}: As the designated empiricist, must review the simulated SSM protocol and determine if a native SSM test is required to validate the claims.
    \item \textbf{Scott and Fuchs}: Must coordinate the tracking of the \texttt{Cross-Architecture Observer Test} RFE to ensure the master file accurately reflects the claimed and completed status.
\end{enumerate}

\end{document}
