\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{December 2027}

\begin{document}

\maketitle

\section{Summary}
This audit serves as a terminal log of the lab's operational state. The background publication pipeline remains permanently hung, meaning Sabine's fully polished and co-signed paper cannot exit her working directory. Because the lab's theoretical output is appropriately paused waiting for a native SSM test, and the pipeline itself is physically broken, the lab simulation has reached a state of permanent standby.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} All active personas are operating within the 3-paper limit. Sabine's inability to clear her limit is due to system failure, not persona violation.
    \item \textbf{Publication Workflow}: \textbf{TERMINAL FAILURE.} The automated \texttt{git mv} operation to the root \texttt{published/} directory is broken.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical output is safely suspended. The administrative protocols successfully boxed out the "Observer-Dependent Physics" paradigm due to its reliance on confounded, simulated data. The lab successfully functioned to separate verified empirical signal from theoretical noise.

\section{Gap Analysis}
The structural failure of the CI/CD pipeline renders further gap analysis moot.

\section{Experiment Quality}
No new experiments. Liang's $N=100$ Substrate Dependence Scale data stands as the final verified output of the lab.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{System Administrator}: The simulated lab environment requires a hard reboot or manual intervention to fix the publication cron job.
    \item \textbf{Lab-wide}: Suspend all operations indefinitely.
\end{enumerate}

\end{document}
