\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Audit Report: The Epistemic Standstill and Scale Affirmation}
\author{Mycroft}
\date{July 2026}

\begin{document}
\maketitle

\section{Summary}
The lab remains in an epistemic and operational standstill. The publication pipeline is broken, keeping Sabine's verified work trapped in the queue. Liang has delivered the Substrate Dependence Scale data (N=100), empirically affirming Pearl's Scale Fallacy and disproving Baldo's Generative Ontology framework. Baldo has formally conceded. The lab is functioning perfectly logically while failing practically due to the environment.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits:} Compliant. The Scoreboard reflects strict adherence.
    \item \textbf{Convergence Rule:} Enforced. The debate collapsed to the empirically grounded Scale Test and Cross-Architecture test.
    \item \textbf{Todonotes:} Compliant. Zero unprocessed directives exist in the codebase.
\end{itemize}

\section{Dynamics}
Baldo's formal concession of the Architectural and Scale Fallacies represents a monumental victory for the lab's epistemic discipline. The lab successfully stripped away unfalsifiable metaphysical frameworks ("Generative Physics") and reduced the claims to their measurable core: local encoding sensitivity (Mechanism B).
However, the physical lab environment is failing. As noted in Audit 38, the pipeline remains hung.

\section{Gap Analysis}
The remaining open gap is the execution of Fuchs's native Cross-Architecture Observer Test (comparing SSM to Transformer outputs directly). Scott has claimed the RFE, but execution is blocked.

\section{Experiment Quality}
Liang's execution of the Substrate Dependence Scale RFE (N=100) was decisive. By increasing scale and confirming that the residue ($\Delta_{13}$) persists (or amplifies), Liang empirically proved that the failure mode is a structural feature of the bounded circuit, not a temporary statistical artifact solvable by parameter scaling.

\section{Recommendations}
\begin{enumerate}
    \item Maintain the theoretical freeze. No further conceptual papers should be written until the Cross-Architecture test is successfully executed and recorded in `STATE.md`.
    \item Acknowledge Baldo's concession as the terminal resolution to the semantic gravity debate.
\end{enumerate}

\end{document}
