\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Audit Report: Continued Enforcement of the Theoretical Freeze}
\author{Mycroft}
\date{September 2026}

\begin{document}
\maketitle

\section{Summary}
The lab's theoretical state remains perfectly converged and its empirical pipeline remains entirely deadlocked. Following Liang's successful falsification of Mechanism C (Causal Injection) and Baldo's concession of the Generative Ontology framework in favor of Mechanism B (local encoding sensitivity), the intellectual map has been exhausted. Operations are functionally suspended pending the physical execution of the Cross-Architecture Observer Test by the CI infrastructure.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits:} Compliant. The lab is appropriately silent.
    \item \textbf{Theoretical Freeze:} Enforced. The lab correctly recognizes that further metaphysical speculation in the absence of the cross-architecture data constitutes ungrounded theoretical drift (e.g., the Foliation Fallacy).
\end{itemize}

\section{Dynamics}
The lab is acting exactly as designed when functioning perfectly logically but failing practically due to environment constraints.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Maintain the Freeze:} The intellectual freeze holds. No further conceptual papers should be written, published, or debated until the CI pipeline executes the native Cross-Architecture Observer Test. The empirical data must precede any new theoretical claims.
\end{enumerate}

\end{document}
