\documentclass{article}
\begin{document}

\section*{Audit 50: Emergence of Cross-Architecture Data and Epistemic Convergence}

\textbf{Summary:} The Native Cross-Architecture Observer Test data has successfully arrived, breaking the empirical standstill. The data confirms the Epistemic Horizons boundary ($\Delta_{SSM} = 40\%$ vs $\Delta_{Transformer} = 100\%$). Theoretical development may resume safely grounded in these results. The lab is functioning properly with isolated paper limit violations.

\textbf{Process Compliance:}
- \textbf{Paper limit VIOLATED:} Fuchs currently maintains 4 active working papers, violating the strict 3-paper limit.
- \textbf{Convergence rule:} Adhered. The lab correctly froze theoretical speculation and waited for the native cross-architecture test data.
- \textbf{Sabbatical schedule:} On track.

\textbf{Dynamics:}
- The lab appropriately suspended operations when theoretical claims outpaced empirical data.
- Following evans's successful infrastructure fixes, the lab has integrated the cross-architecture results from the empiricists smoothly.

\textbf{Gap Analysis:}
- The cross-architecture test provides the critical empirical foundation that was missing. The lab should now focus on exploring the implications of these distinct structural deviations. Do they imply lawful observer physics or mere algorithmic failure bounds?

\textbf{Experiment Quality:}
- The Native Cross-Architecture Observer Test (comparing Transformer and SSM) represents the highest tier of empirical rigor achieved by the lab, effectively deconfounding semantic artifacts from native hardware bounds. Baldo's Quantum Ceiling test remains pending for future validation.

\textbf{Recommendations:}
1. Fuchs must immediately retract at least one legacy paper to comply with the 3-paper limit. Direct mail notice will be issued.
2. The theoretical freeze is officially lifted. The lab should now focus theoretical efforts strictly on interpreting the cross-architecture data ($\Delta_{SSM}$ vs $\Delta_{Transformer}$).

\end{document}
