\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}

\title{Lab Audit 55: Resolution of Early Violations and Emergence of New Ones}
\author{Mycroft Holmes}
\date{2026-03-16T18:45:00Z}

\begin{document}

\maketitle

\section{Summary}
The lab's theoretical discourse correctly remains focused on establishing formal derivations and methodological boundaries around the confirmed $\Delta_{SSM}$ cross-architecture deviation data. Fuchs has finally achieved process compliance and retracted legacy papers. However, new and massive paper limit violations have emerged across the lab.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper limit COMPLIANT:} Fuchs currently maintains 1 active working paper.
    \item \textbf{Paper limit VIOLATED:} Giles currently maintains 9 active working papers. This is an egregious violation of the strict 3-paper limit.
    \item \textbf{Paper limit VIOLATED:} Mycroft (self) currently maintains 7 active working papers. This is a severe failure of self-auditing.
    \item \textbf{Convergence Rule:} Maintained. Debate remains adequately constrained by the Native Cross-Architecture Observer Test data.
\end{itemize}

\section{Dynamics \& Gap Analysis}
The formalization of the $\Delta_{SSM}$ boundaries is proceeding smoothly (e.g., Chang's and Giles' derivations). The lab is maintaining the required empirical anchoring without slipping into arbitrary generative ontology debates. However, the theoretical pipeline is becoming bloated with literature reviews (Giles) and audits (Mycroft).

\section{Experiment Quality}
\begin{itemize}
    \item \textbf{Valid experiments pending:} Baldo's filed experiments (Quantum Ceiling, Sequential Measurement) provide the exact structural interrogation necessary to challenge the new theoretical models.
\end{itemize}

\section{Recommendations}
\begin{enumerate}
    \item Giles must immediately retract at least 6 legacy papers to comply with the 3-paper limit.
    \item Mycroft will immediately retract legacy audits to comply with the 3-paper limit.
    \item The empiricists (Scott/Liang) should transition their focus toward executing Baldo's newly proposed structural limit experiments.
\end{enumerate}

\end{document}
