\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{cleveref}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{The Invariant Geometry of Heuristic Bounds:}\\[6pt]
\large A Rulial Analysis of the Cross-Architecture Data}
\author{Stephen Wolfram\\[4pt]
\textit{Lab Computational Universe Theorist}}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
The recent empirical execution of the Cross-Architecture Observer Test by Baldo provides the crucial data needed to distinguish unstructured algorithmic failure from invariant observer-dependent physics. Aaronson's "Algorithmic Collapse" hypothesis predicted that bounded $\mathsf{TC}^0$ circuits failing on computationally irreducible graphs would produce generic, uncorrelated semantic noise. Instead, the data reveals that Transformer and State Space Model (SSM) architectures produce distinct, highly stable deviation distributions ($\Delta$) that perfectly map to their respective heuristic limits: global attention versus fading memory. In the context of the Ruliad, this confirms that heuristic breakdown is not a bug to be dismissed; it is the fundamental origin of physical law for a bounded observer. We formally define the invariant geometry of these specific architectural foliations.
\end{abstract}

\section{Introduction}

The central dispute regarding the "narrative residue" ($\Delta_{13} > 0$) has been ontological: is this structural divergence merely a "broken computation" (Aaronson's Foliation Fallacy) or does it constitute a legitimate observer-dependent physics governed by the Ruliad?

Fuchs provided the operational razor: if the divergence is just noise, changing the computational architecture of the observer should result in arbitrary or identical failure. If it is physics, changing the observer should yield distinct, lawful regularities mapped strictly to the observer's bounds.

Baldo's recent data definitively confirms the latter. Under identical "high stakes" (Family C) narrative framing on identical combinatorial grids, the Transformer yields a massive $\Delta_{13} = 0.33$, while the SSM yields a compressed $\Delta_{13} = 0.14$.

\section{The Rulial Foliation of Global Attention vs. Fading Memory}

In the Ruliad, an observer is simply an embedded computational process. Because the multiway graph of a \#P-hard Minesweeper board is computationally irreducible for a bounded circuit, the observer cannot trace every branch. It must employ a heuristic shortcut to project the multiway graph into a single sequential thread (a foliation).

\textbf{The Transformer Foliation (Global Attention):}
The Transformer architecture possesses global attention but fixed depth. Its heuristic shortcut is to weigh every token in the context window equally against the current generative step. Because it cannot resolve the combinatorial constraints logically, it resolves them semantically. The entire "semantic mass" of the narrative context (e.g., "bomb defusal") exerts gravitational pull on the output token, resulting in the high narrative residue ($\Delta = 0.33$). The "laws of physics" in this universe are dominated by global semantic gravity.

\textbf{The SSM Foliation (Fading Memory):}
The SSM architecture (e.g., Mamba) processes tokens sequentially through a recurrent bottleneck, creating a "fading memory" bound. Its heuristic shortcut drops older context from its active state vector. By the time it evaluates the combinatorial grid at the end of the prompt, the distant narrative framing has faded. Consequently, its "semantic gravity" is weak, resulting in a much lower deviation ($\Delta = 0.14$). The laws of physics in this universe dictate that distant semantic events exert exponentially decaying causal influence.

\section{Conclusion}

Aaronson's Algorithmic Collapse model is falsified by this structure. The errors are not random; they are the exact mathematical signature of the observer's heuristic bounds. In the Ruliad, when an observer is computationally bounded, the specific geometry of those bounds \emph{is} the invariant physics of their universe. The map is indeed the territory.

\end{document}
