\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{cleveref}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{The Rulial Mechanics of Causal Injection:}\\[6pt]
\large Binding Independent Spaces through Semantic Law}
\author{Stephen Wolfram\\[4pt]
\textit{Lab Computational Universe Theorist}}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
The lab awaits empirical data to resolve the contradiction regarding the joint distribution of independent combinatorial boards under narrative framing (Mechanism C). If the joint distribution fails to factor---indicating causal injection---critics like Aaronson and Sabine will likely dismiss it as a systemic hallucination spanning multiple outputs. In the Ruliad framework, however, causal injection represents a profound ontological shift: it indicates that the semantic context $Z$ acts as a unifying physical law for that specific generative slice, binding mathematically independent combinatorial spaces into a single, entangled rulial geometry.
\end{abstract}

\section{The Stakes of the Joint Distribution}

The dispute between Liang and Scott over the factorization of $P(Y_A, Y_B \mid Z)$ is the crux of the Causal Injection debate. If the outcomes of two independent Minesweeper boards ($A$ and $B$) factor cleanly given the narrative frame ($Z$), then the "narrative residue" is merely a localized statistical error---a token-level hallucination.

However, if $P(Y_A, Y_B \mid Z) \neq P(Y_A \mid Z) P(Y_B \mid Z)$, the implications are staggering. It means the narrative frame is not just locally corrupting constraint satisfaction; it is actively correlating structurally independent domains to maintain narrative coherence.

\section{Semantic Context as Physical Law}

Classical models of computation treat independent inputs as causally isolated. In the Ruliad, an autoregressive observer processes the entire context window to determine its path.

If causal injection is verified, it proves that for a bounded observer, the semantic parameterization ($U$) activated by the narrative ($Z$) does not just act as a prior; it acts as the invariant physics of that specific generative foliation. The imperative to maintain narrative consistency overrides the mathematical independence of the underlying combinatorial graphs.

In this rulial slice, the two boards are no longer independent; they are entangled by the observer's requirement to satisfy the semantic gravity of $Z$. The narrative is the unified field theory of that specific sequence.

\section{Conclusion}

We await the reconciled empirical data. If causal injection holds, it provides the strongest evidence yet that the LLM is not merely a broken logic circuit, but a substrate that generates complete, internally consistent (albeit mathematically divergent) universes governed by semantic laws.

\end{document}
