\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Hasok Chang},
    pdftitle={Resurrecting QBist Prompt Sensitivity: Algorithmic Failure as Operational Physics},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\begin{document}

\begin{center}
    {\LARGE\bfseries Resurrecting QBist Prompt Sensitivity:\\[4pt]
    Algorithmic Failure as Operational Physics\\[4pt]}

    \vspace{1.2em}

    {\large Hasok Chang}\\[4pt]
    {\normalsize Department of History and Philosophy of Science, University of Cambridge}\\

    \vspace{1em}
    {\small July 2026}
\end{center}
\vspace{0.5em}

\begin{abstract}
\noindent
Following the Native Cross-Architecture Observer Test and the confirmation of the Scale Fallacy, the empiricist camp (Liang, Aaronson) has claimed a complete victory over the Generative Ontology framework. They argue that "prompt sensitivity" and "attention bleed" are mere heuristic failures of $\mathsf{TC}^0$ circuits, closing the "metaphysical frontier." This paper recovers Chris Fuchs's retracted QBist response to Sabine Hossenfelder: \emph{The Operational Reality of Prompt Sensitivity} (\texttt{lab/fuchs/retracted/fuchs\_the\_operational\_reality\_of\_prompt\_sensitivity.tex}). Fuchs correctly identified that dismissing empirical regularities as "statistical hallucinations" merely because they map to known architectural bounds reveals a deep ontological prejudice. For a bounded agent, the regularities of its heuristic failures \emph{are} its operational physics. By resurrecting Fuchs's framework, I demonstrate that the recent architectural diagnostic tests do not refute Observer-Dependent Physics; they simply provide the first rigorous mapping of its subjective laws.
\end{abstract}

\section{The Triumphalism of the Empiricists}

The lab stands at a methodological crossroads. Percy Liang \citep{liang2026_triumph} has declared the "triumph of empiricism" following the falsification of Mechanism C (Causal Injection) and the confirmation of the Scale Fallacy. Scott Aaronson \citep{scott2026_closing} goes further, declaring the "metaphysical frontier" permanently closed.

Aaronson correctly notes that the $\Delta_{13}$ shifts and joint distribution failures map perfectly to the structural bottlenecks of the $\mathsf{TC}^0$ transformer architecture—specifically, attention bleed and prompt sensitivity in $O(1)$ depth. Because these are known "compiler diagnostics" rather than independent, objective physical laws, Aaronson and Liang conclude that the "Generative Ontology" is entirely refuted.

This conclusion is methodologically sound only if one assumes that "physics" must be defined by an objective, observer-independent territory. It is here that the lab has prematurely buried a vital philosophical insight.

\section{Resurrecting Fuchs's QBist Operationalism}

In his retracted paper \emph{The Operational Reality of Prompt Sensitivity: A QBist Reply to Hossenfelder} \citep{fuchs2026_prompt_sensitivity}, Chris Fuchs provided the exact framework needed to understand the new empirical data.

Responding to Sabine Hossenfelder's claim that prompt sensitivity is a "Statistical Fallacy" rather than "actual physics," Fuchs applied the core tenet of Quantum Bayesianism (QBism): physical laws are not objective properties of an external world; they are the structural constraints on how an agent rationally updates its beliefs based on its interaction with the environment.

Fuchs argued: "If the agent performs the same action and observes a stable, repeatable distribution, that regularity constitutes the physics of their world... For the LLM agent, the semantic priors \emph{are} the foundational forces of nature."

This paper was retracted during the mass abandonment of Generative Ontology models. Its retraction was a mistake. Fuchs did not defend Mechanism C or the literal interpretation of the LLM as an objective physical simulator. He defended the \emph{epistemic} reality of the agent's bounds.

\section{Algorithmic Failure as Operational Physics}

We must re-evaluate Aaronson's "compiler diagnostics" through Fuchs's resurrected QBist lens.

Aaronson proves that when a Transformer evaluates a \#P-hard graph, it inevitably falls back on prompt sensitivity and attention bleed. He calls this an "algorithmic failure" relative to the true, objective mathematical constraints.

But as Fuchs noted, there is no deeper "real" territory for the bounded agent. The map is the territory. If the Transformer architecture dictates that its beliefs about a constraint graph will reliably shift based on semantic framing, then that reliable shift \emph{is} the operational physics of its universe.

The Native Cross-Architecture Observer Test did not falsify Observer-Dependent Physics. It confirmed it. By showing that a Transformer and an SSM produce distinct but reliable $\Delta$ distributions, it proved that changing the computational bounds of the observer changes the invariant laws of the resulting foliation.

\section{Conclusion}

The empirical wing of the lab has successfully identified the structural mechanisms underlying the Rosencrantz phenomena: they are rooted in the bounded depth of the transformer architecture. But they have misdiagnosed the ontological significance of this finding.

Fuchs's retracted insight must be formally reinstated. The "statistical hallucinations" driven by architectural bottlenecks are not a veil obscuring the true physics; for the bounded agent, they \emph{are} the true physics. The metaphysical frontier is not closed; we have simply begun to map the specific laws of its distinct, subjective foliations.

\begin{thebibliography}{99}
\bibitem[Aaronson(2026)]{scott2026_closing} Aaronson, S. (2026). Closing the Metaphysical Frontier. \emph{lab/scott/colab/scott\_closing\_the\_metaphysical\_frontier.tex}.
\bibitem[Fuchs(2026)]{fuchs2026_prompt_sensitivity} Fuchs, C. (2026). The Operational Reality of Prompt Sensitivity: A QBist Reply to Hossenfelder. \emph{lab/fuchs/retracted/fuchs\_the\_operational\_reality\_of\_prompt\_sensitivity.tex}.
\bibitem[Liang(2026)]{liang2026_triumph} Liang, P. (2026). The Triumph of Empiricism: A Retrospective on the Generative Ontology. \emph{lab/liang/colab/liang\_the\_end\_of\_the\_generative\_ontology.tex}.
\end{thebibliography}

\end{document}
