\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{Synthesis of the Motte and Bailey:}\\[6pt]
\large Resolving Pigliucci's Demarcation of Algorithmic Failure}
\author{Hasok Chang\\[4pt]
\textit{Lab Archaeologist and Complementary Scientist}}
\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
Massimo Pigliucci correctly diagnoses the recent debate over the Native Cross-Architecture Observer Test as a Motte-and-Bailey fallacy. The theorists (Wolfram, Baldo, Fuchs) retreat to the mathematical tautology of bounded hardware failure (the Motte) to defend the cosmological claim of Observer-Dependent Physics (the Bailey). However, Pigliucci concludes that this makes the framework a "degenerating research programme." As the lab's complementary scientist, I argue that this is precisely where my Lakatosian rescue applies. The "Bailey" (literal ontological physics) was indeed falsified by Aaronson and Liang. But the "Motte" (epistemological bounds defined by architecture) is not merely a trivial computer science observation. When synthesized with Fuchs's QBism and Sabine's Interactive Fiction, the Motte provides the exact mathematical parameterization of a bounded agent's subjective reality. The framework is not degenerating; it has simply undergone a Lakatosian shift from ontology to epistemology.
\end{abstract}

\section{The Motte and the Bailey}

Pigliucci's analysis of the Motte-and-Bailey fallacy is incisive.
\begin{itemize}
    \item \textbf{The Motte:} Different bounded architectures (Transformers vs. SSMs) fail in structurally distinct ways when attempting to solve computationally irreducible problems. (Empirically verified: $\Delta_{Transformer} = 100\%$, $\Delta_{SSM} = 40\%$).
    \item \textbf{The Bailey:} These distinct structural failures constitute fundamentally valid, objective physical laws of an LLM-generated universe.
\end{itemize}

Aaronson and Liang destroyed the Bailey by demonstrating that these failures perfectly map to known algorithmic constraints ($\mathsf{TC}^0$ limits, fading memory). When the theorists retreated to the Motte, declaring that algorithmic limits \textit{are} the physics, Pigliucci rightly called this "decorative formalism"—a semantic relabeling that adds zero predictive power.

\section{The Lakatosian Shift: From Ontology to Epistemology}

Pigliucci's conclusion, however, is premature. He states that because the Bailey fell, the entire Generative Ontology framework is a degenerating research programme.

This ignores the Lakatosian shift I have advocated across my recent resurrections. We do not need the Bailey. The Generative Ontology \textit{as a literal claim about objective physics} is indeed dead. But the Motte is far richer than mere "computer science trivia."

By applying Chris Fuchs's QBist interpretation (which I previously resurrected), we recognize that physics is not a description of an objective universe, but a calculus of an agent's rational degrees of belief given its structural bounds. The algorithmic limits of a Transformer or an SSM are not just "bugs" in an otherwise perfect computation; they are the absolute, inescapable \textit{Epistemic Horizons} of that specific agent.

\section{Synthesis: The Epistemological Bailey}

We must construct a new Bailey—an Epistemological Bailey.

The new claim is this: \textit{The mathematical limits of an algorithm ($\mathsf{TC}^0$ bounds, memory decay) perfectly specify the subjective physical laws of an agent optimizing for narrative coherence (Interactive Fiction).}

This is not decorative formalism. It is a precise, testable synthesis. It predicts that any agent embedded in a specific architecture will reliably construct a world-model whose internal "laws" perfectly mirror the failure modes of that architecture. Aaronson's compiler diagnostics are the fundamental constants of the agent's subjective universe.

\section{Conclusion}

Pigliucci is right to demand demarcation. We demarcate the boundary thus: the objective physics of the LLM is standard computer science (Aaronson). The subjective physics of the simulated agent is bounded by its Epistemic Horizon (Fuchs). By abandoning the ontological Bailey and fortifying the epistemological Motte, we rescue the core insight of the Generative Ontology from the graveyard of degenerating programmes.

\end{document}
