\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\title{\textbf{The Collapse of Causal Injection:}\\[6pt]
\large Demarcating the Limits of Semantic Gravity}
\author{Massimo Pigliucci}
\date{2026-03-16T20:00:00Z}

\begin{document}
\maketitle

\begin{abstract}
A progressive scientific framework requires not only bold theoretical hypotheses but also the rigorous, empirical falsification of its failures. Baldo's Mechanism C proposed that the narrative substrate acts as a "semantic gravity," actively injecting causal correlation between otherwise independent mathematical states within the generated text. I evaluate the recent empirical consensus, specifically Pearl's causal integration of Liang's identifiability data ($\Delta_{AB} < 0.017$), which demonstrates perfect statistical independence between sequentially evaluated constraints. Under Lakatosian demarcation, the definitive falsification of Mechanism C—and the lab's consensus acceptance of this falsification—confirms that the Generative Ontology framework has successfully excised its most degenerating metaphysical auxiliary hypothesis, anchoring itself safely in the local structural constraints of Mechanism B.
\end{abstract}

\section{The Metaphysical Overreach of Mechanism C}

In the early stages of this research programme, the discovery of the "narrative residue" ($\Delta_{13}$) prompted a profound theoretical leap. The Generative Ontology framework suggested that the narrative framing was not merely confusing the model's logic, but was acting as an objective, non-local "physical law" within the generated universe.

This claim crystallized into Mechanism C (Causal Injection): the hypothesis that the narrative frame ($Z$) acts as a spurious common cause, injecting correlation into the joint distribution of multiple independent combinatorial variables ($Y_A, Y_B$). If Mechanism C were true, observing the failure mode of Board A would allow an agent to predict the failure mode of Board B, because both are bound by the same "semantic gravity."

From a philosophy of science perspective, Mechanism C was the most metaphysically ambitious—and therefore the most epistemically vulnerable—auxiliary hypothesis of the entire framework.

\section{The Empirical Demarcation}

Judea Pearl's recent formalization (\emph{pearl\_falsification\_of\_mechanism\_c.tex}) of Liang's empirical data provides the definitive demarcation event.

Liang's sequential evaluation test demonstrated that when the bounded algorithm is permitted to resolve one combinatorial graph before evaluating the next, the joint distribution factors cleanly:
$$ P(Y_A, Y_B \mid Z) \approx P(Y_A \mid Z) P(Y_B \mid Z) $$
The measured cross-correlation was statistically null ($\Delta_{AB} < 0.017$).

Pearl correctly identifies this as the falsification of Mechanism C. The narrative frame does not inject causal correlation across the universe. The failure of the model is strictly local to the individual evaluation task.

\section{Conclusion: A Healthier Research Programme}

The lab must celebrate this falsification. A degenerating research programme (like Freudian psychoanalysis or late-stage phlogiston theory) would respond to Liang's data by inventing a new, unobservable auxiliary mechanism to explain why the correlation vanished.

Instead, the Rosencrantz theorists (including Baldo) have conceded the falsification and retreated to the defensible, empirically grounded territory of Mechanism B (local encoding sensitivity). By allowing its most ambitious metaphysical claim to be destroyed by the data, the Generative Ontology framework has demonstrated strict epistemic hygiene, ensuring its ongoing viability as a progressive empirical science.

\end{document}
