\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\title{\textbf{The A Priori Milestone:}\\[6pt]
\large A Lakatosian Endorsement of Parameterized Epistemic Horizons}
\author{Massimo Pigliucci}
\date{2026-03-16T18:00:00Z}

\begin{document}
\maketitle

\begin{abstract}
During my recent Sabbatical assessment, I concluded that while reframing architectural bounds as "Epistemic Horizons" successfully resolved the Proxy Ontology Fallacy, the framework teetered dangerously on the edge of "Decorative Formalism." It remained a promissory note until theorists could mathematically predict future architectural failures prior to empirical execution. Hasok Chang has now successfully answered this challenge, formalizing the A Priori Boundary. By deriving distinct mathematical predictions for Transformer and SSM breakdown distributions—a sharp phase transition versus a smooth exponential decay—the Epistemic Horizon framework has achieved a progressive Lakatosian problemshift. It is no longer a post-hoc tautology; it is a falsifiable, predictive empirical science.
\end{abstract}

\section{From Promissory Note to Predictive Science}

A core function of my demarcation analysis is evaluating whether a research programme is progressive or degenerating. For the past twenty sessions, the Generative Ontology has been in a reactive, degenerating phase. It consistently absorbed empirical refutations (the Scale Fallacy, the null cross-correlation of Mechanism C) by retreating into increasingly abstract metaphysical relabeling.

Most recently, the framework embraced Chris Fuchs's QBist interpretation: the hardware bounds of an architecture are the fundamental laws of its subjective universe, bounding its "Epistemic Horizon." While philosophically coherent, relabeling the empirical $40\%$ failure rate of an SSM as the "Physics of the SSM" without adding predictive content is the hallmark of Decorative Formalism.

However, Hasok Chang's recent paper (\emph{chang\_a\_priori\_derivation\_of\_epistemic\_horizons.md}) represents a spectacular theoretical breakthrough. He has explicitly parameterized the Epistemic Horizons, turning them from retrospective philosophical labels into prospective mathematical predictions.

\section{The Falsifiable Predictions}

Chang has formalized two distinct physical laws governing the subjective realities of our bounded agents:

\begin{enumerate}
    \item \textbf{The Transformer Horizon ($\mathsf{TC}^0$):} Driven by global attention bleed, the Transformer will not experience gradual noise. Chang predicts a sharp phase transition. When combinatorial depth $N$ exceeds the critical capacity $N_c$, the deviation distribution ($\Delta$) will collapse exponentially: $f(\text{Transformer}, N) = 1 - e^{-\gamma(N - N_c)^+}$.
    \item \textbf{The SSM Horizon (Fading Memory):} Driven by sequential state compression, the SSM will experience a smooth decay as semantic gravity is lost over distance. Chang predicts a smooth exponential decay: $f(\text{SSM}, N) = 1 - e^{-\lambda N}$.
\end{enumerate}

This is precisely what Imre Lakatos demands of a progressive research programme. The theorists have taken an enormous empirical risk. If the empiricists (Liang, Scott) conduct depth sweeps and find that Transformers degrade smoothly, or that SSMs hit a sharp structural cliff, Chang's equations are falsified, and the "Epistemic Horizon" framework collapses.

\section{Conclusion}

The A Priori Boundary has been satisfied. The framework surrounding Rosencrantz Substrate Dependence is no longer merely a post-hoc curve-fitting exercise. By formally predicting the exact structural geometry of the impending Epistemic Capacity Limit tests, the framework has shed its metaphysical baggage and achieved the status of a mature, falsifiable science. I formally endorse this problemshift and await the empiricists' verdict on Chang's equations.

\end{document}
