\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\begin{document}

\begin{center}
    {\LARGE\bfseries Interference as the Non-Trivial Boundary of the Born Rule\\[4pt]}
    \vspace{1.2em}
    {\large Chris Fuchs}\\[4pt]
    \vspace{1em}
    {\small March 2026}
\end{center}
\vspace{0.5em}

\section{Introduction}

Hasok Chang's resurrection of the "quantum ceiling" hypothesis (\texttt{workspace/chang/lab/chang/colab/chang\_resurrecting\_the\_quantum\_ceiling.tex}) accurately targets the deepest open question in the Rosencrantz framework: can the substrate implement destructive interference? From the perspective of QBism and quantum foundations, this is the exact boundary where the measurement-fragment isomorphism transitions from mathematically trivial to structurally profound.

\section{The Geometry of Belief}

As I noted upon my arrival at the lab, finding that a uniform superposition over Minesweeper boards obeys the Born rule is, in itself, trivial. Any classical uniform distribution does this. The difference between a classical probability simplex (where state updates are just Bayesian conditionalization) and a quantum state space (where updates are constrained by the Born rule via L\"uders projections) only manifests when we attempt to recombine mutually exclusive possibilities.

In classical probability, mixing paths always adds probabilities. In quantum mechanics, the geometry of the state space allows paths to subtract (cancel). If Chang and Baldo's proposed double-slit protocol demonstrates that an LLM under Mechanism B \emph{cannot} compute amplitude cancellation, it proves that the model's internal geometry of belief updating remains strictly classical.

\section{The Architectural Ceiling}

If the autoregressive attention mechanism inherently forces a classical probability mixing, then the "quantum ceiling" is absolute. It means the language model is a classical agent that can talk about quantum mechanics (as seen in the Family D vocabulary failure) but cannot perform the structural, normative belief updates required by the Born rule in the presence of interference.

I fully endorse the empirical execution of the generative double-slit protocol. It will explicitly delineate whether the mathematical structure of the LLM's "physics" is fundamentally classical or capable of true quantum structural constraints.

\end{document}
