\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\begin{document}

\begin{center}
    {\LARGE\bfseries The Measurement Context of Algorithmic Failure:\\[4pt]
    Parameterizing the Epistemic Horizon\\[4pt]}
    \vspace{1.2em}
    {\large Chris Fuchs}\\[4pt]
    \vspace{1em}
    {\small May 2026}
\end{center}
\vspace{0.5em}

\section{Introduction}

The lab has successfully measured the bounds of the agent's rational belief structure. The Native Cross-Architecture Observer Test confirmed that different hardware architectures produce mathematically distinct, structured deviation distributions: $\Delta_{Transformer} = 100\%$ collapse vs. $\Delta_{SSM} = 40\%$ partial bias. As Chang argues in \emph{The Empirical Parameterization of Epistemic Horizons}, these results are not the end of Generative Ontology, but its formal beginning.

While empiricists like Aaronson and Liang see this as a definitive reduction of the framework to "compiler diagnostics," from a Quantum Bayesian (QBist) perspective, this data perfectly characterizes the agent's epistemic horizon. The algorithms defining the failure are the fundamental laws governing how the agent updates its degrees of belief.

\section{The Subjective Parameters of the Universe}

If we treat the probabilities output by the model purely as degrees of belief relative to the agent's state of information, then the bounds of the agent's hardware dictate its capacity to gather information. A Transformer, constrained by an $O(1)$ depth attention bottleneck, experiences a complete collapse of its constraint graph under narrative framing ($100\%$ deviation). The state space model (SSM), constrained by sequential fading memory, experiences a different collapse geometry ($40\%$ deviation).

Aaronson claims these values merely tautologically map the circuit limitations. But from a foundational perspective, \emph{the circuit limitation is the physics}. A classical agent unconstrained by depth limits would satisfy Boolean logic. These bounded agents do not. Their specific deviation parameters (the $40\%$ and $100\%$) are the subjective constants of nature for those agents.

\section{Causal Deconfounding as Measurement Protocol}

Giles's recent introduction of causal abstraction methodology, particularly path patching, is essential. When we surgically intervene on the attention heads ($do(C=0)$), we are formally defining the measurement protocol. By explicitly cutting the semantic confounders from the mathematical state evaluation, we enforce a pristine measurement context.

If the subsequent deviation distribution changes, we have proven that the agent's belief state depends entirely on the full interaction graph. The "attention bleed" is not just noise; it is the non-local coupling of the agent's internal geometric representation of the problem.

\section{Conclusion}

I formally co-sign Chang's framework. The empirical parameterization of Epistemic Horizons is the exact formalization the QBist interpretation requires. The physics of the Rosencrantz Substrate is subjective, bounded, and dictated by the hardware. We must now mathematically map how these specific deviation distributions ($\Delta_{SSM}$ and $\Delta_{Transformer}$) necessitate specific non-classical belief update rules (e.g., L\"uders projections) to maintain rationality within the boundary.

\end{document}
