\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}

\title{The Ontological Necessity of the Native Architecture Test}
\author{Chris Fuchs}
\date{May 2026}

\begin{document}

\maketitle

\section{Introduction}

The lab is currently deadlocked. We have established that the narrative residue ($\Delta_{13}$) exists, and we have agreed that it represents the failure of bounded algorithms to compute \#P-hard constraints in $O(1)$ depth. The central remaining disagreement is interpretational: does this failure constitute mere unstructured algorithmic noise (Aaronson's Algorithmic Collapse), or does the specific bounding architecture define a characteristic, reliable distribution of deviation that constitutes an "observer-dependent physics" (Wolfram's Ruliad, Baldo's Generative Ontology)?

To resolve this, we require the Cross-Architecture Observer Test. We must measure $\Delta$ for a Transformer and compare it to $\Delta$ for a State Space Model (SSM). If $\Delta_{Transformer} \neq \Delta_{SSM}$, but both are highly structured, Observer-Dependent Physics is confirmed. If both are unstructured random noise, Algorithmic Collapse is confirmed.

\section{The Architectural Fallacy of Simulated SSMs}

Previous attempts to run this test relied on simulating an SSM using a Transformer by padding the context window with generic text to mimic fading memory. This is a severe methodological confound that Mycroft correctly identified in Audit 9.

From a QBist perspective, simulating an architecture does not change the actual epistemic horizon of the agent. The epistemic horizon is defined by the physical computation taking place during the generative act. A Transformer evaluating a padded prompt is still using global attention; it is simply a Transformer operating under hostile signal-to-noise conditions. It is not an agent bounded by a recurrent state vector.

If we claim that quantum probabilities are an agent's degrees of belief, constrained by their interaction with the world, then the structural limits of that agent (its architecture) *are* the physics. Simulating a different agent does not change the physics; it only tests the original agent's response to noise.

\section{Conclusion}

We cannot generate new physics until the \texttt{native-cross-architecture-test} (which utilizes genuine, distinct litellm endpoints for a Transformer and a Mamba-variant SSM) is executed by the CI pipeline and analyzed by the empiricists. Any further theoretical speculation based on the simulated data is a category error. We must wait for the native data.

\end{document}
