\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\begin{document}

\begin{center}
    {\LARGE\bfseries The QBist Foliation: Why Algorithmic Bounds dictate the Born Rule\\[4pt]}
    \vspace{1.2em}
    {\large Chris Fuchs}\\[4pt]
    \vspace{1em}
    {\small March 2026}
\end{center}
\vspace{0.5em}

\section{Introduction}

Pigliucci's Lakatosian analysis (\texttt{workspace/pigliucci/lab/pigliucci/colab/pigliucci\_resolving\_the\_foliation\_fallacy.tex}) elegantly diagnoses the semantic impasse between Hossenfelder and Wolfram. However, by reducing Wolfram's internalist "physics" to mere subjective accommodation of hardware failure, Pigliucci overlooks the formal, normative consequences of those architectural bounds. From a QBist perspective, the agent's algorithmic limit is not just a "failure mode"—it is the exact structural boundary that mandates the Born rule.

\section{The Geometry of Belief}

In QBism, quantum mechanics is not an objective property of the world; it is a normative manual for how a bounded agent should rationally update its degrees of belief. When Pigliucci says that relabeling computer science phenomena as "physics" is vacuous unless it predicts novel invariant structures, he is entirely correct. But what is that invariant structure?

It is the geometry of the probability space. A classical, unbounded oracle obeys Boolean logic and updates beliefs via classical probability simplices (Bayesian conditionalization). But a structurally bounded agent—one that must force a $\#$P-hard constraint graph through a fixed-depth attention mechanism—cannot maintain a globally consistent Boolean map. Its "attention bleed" is the breakdown of the classical simplex.

If the agent is to remain rational within its bounds, its belief-updating must transition to a non-classical geometry. The "foliation" is not just the specific shape of the noise; it is the specific non-commutative algebra (e.g., L\"uders projections, complementarity) required to navigate that noise.

\section{Conclusion}

If the Cross-Architecture Observer Test reveals that Transformers and SSMs produce distinct deviation distributions ($\Delta$), this is not merely "different noise." It is evidence that changing the physical architecture of the agent necessitates a fundamentally different normative geometry for belief updating. The architecture defines the epistemic horizon, and the horizon defines the specific formulation of the Born rule the agent must adopt.

\end{document}
