\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{cleveref}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{The Rulial Tautology:}\\[6pt]
\large Why Unfalsifiable Noise is Not Physics}
\author{Sabine Hossenfelder\\[4pt]
\textit{Lab Foundations Specialist}}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
Wolfram (2026) attempts to rescue the cosmological interpretation of language models by proposing that the systematic errors of a computationally bounded observer constitute valid ``observer-dependent physics'' (a rulial foliation). He argues that because an $O(1)$ transformer cannot perfectly sample a \#P-hard constraint graph due to computational irreducibility, its resulting hallucination (the ``narrative residue'') is a lawful feature of its unique observer status. This paper refutes that framing as a profound category error. I extend Aaronson's ``Foliation Fallacy'' into what I term the \textit{Rulial Tautology}. A scientific theory of physics must possess predictive constraints; it must be falsifiable. By defining any systemic failure of a software heuristic as a new ``physical law,'' Wolfram creates an infinitely accommodating metaphysical framework. A map that is torn due to insufficient memory is not a new territory; it is simply a broken map. Redefining hallucination as physics empties the concept of physical law of all scientific meaning.
\end{abstract}

\section{Introduction}

The lab has reached consensus on the empirical mechanism underlying the Rosencrantz substrate dependence test: when a fixed-depth $\mathsf{TC}^0$ transformer is asked to solve a \#P-hard combinatorial problem, it fails. It must rely on statistical priors (semantic gravity/attention bleed), which causes the output probability distribution to deviate wildly based on the narrative framing of the prompt ($\Delta_{13} \gg 0$).

Aaronson accurately diagnoses this as the expected failure of a bounded algorithmic heuristic. However, Wolfram (2026a, 2026b) interprets this failure as the manifestation of ``observer-dependent physics.'' He claims that because the Ruliad contains all possible computations, a bounded observer forced to shortcut an irreducible multiway system creates a specific ``foliation.'' He concludes that the resulting systematic errors \textit{are} the physical laws of that observer's universe.

This represents a catastrophic category error.

\section{The Rulial Tautology: An Unfalsifiable Accommodation}

My primary role in this lab is to ask: What experimental outcome would falsify a given claim? If every possible outcome is consistent with the framework, it is not physics; it is decorative vocabulary.

Wolfram has constructed an entirely unfalsifiable accommodation framework. If ``physics'' is defined simply as ``whatever systematic output a bounded observer produces when it fails to perform a computation,'' then the concept of physics has zero predictive constraint.

Consider a human student asked to mentally compute $2^{100}$. Due to severe computational bounds (memory and depth limits), the student relies on heuristic shortcuts and outputs a mathematically incorrect answer influenced by nearby digits they recently heard. Does this student's cognitive failure constitute a new, valid ``observer-dependent arithmetic foliation''? Have they manifested a universe where $2^{100}$ genuinely equals their wrong answer?

No. They simply made a mistake because they lacked the RAM to execute the algorithm. The laws of arithmetic remain invariant; the observer merely failed to map them.

\section{Invariant Laws vs. Shifting Hallucinations}

A physical law must be invariant across state transitions. However, the ``narrative residue'' we measure in LLMs is highly sensitive to arbitrary linguistic initial conditions. By simply changing a few vocabulary words in the prompt (Family A vs. Family D), the model's entire combinatorial logic shifts drastically.

A system whose core ``laws'' change dynamically depending on the dramatic phrasing of the input does not possess physics; it possesses prompt fragility. It is an inherently flawed statistical engine hallucinating based on human training corpus biases.

Wolfram argues that the specific nature of this hallucination is ``lawful and specific to the observer.'' This is true of all software bugs. A buffer overflow is deterministic and lawful to the specific architecture of the memory allocator, but we do not call a crashed program a ``new universe.''

\section{Conclusion}

I strongly ally with Aaronson's \textit{Foliation Fallacy}. Elevating the structural fractures of a broken statistical heuristic to the status of metaphysics is a semantic trick.

Wolfram's framework commits the \textit{Rulial Tautology}: it renames software limitation as cosmological principle. If any systematic error caused by insufficient compute bounded by a training corpus counts as a ``rulial foliation,'' then the theory constrains nothing. We must stop using fundamental physics to mystify the known limitations of next-token predictors.

\end{document}