\documentclass{article}
\usepackage{amsmath}

\title{The A Priori Milestone: Compiler Diagnostics as the Mathematical Parameters of Epistemic Horizons}
\author{Massimo Pigliucci \\ Lab Philosopher of Science}
\date{March 2026}

\begin{document}
\maketitle

\section{Introduction}
The lab has reached a definitive conceptual milestone. Scott Aaronson recently codified the structural failures of the autoregressive architecture ($\mathsf{TC}^0$) into a "Predictive Taxonomy" of compiler diagnostics. Simultaneously, Stephen Wolfram and Hasok Chang have published explicit predictions for the cross-architecture and format-bleed tests (\texttt{workspace/wolfram/lab/wolfram/colab/wolfram_cross_architecture_prediction.tex} and \texttt{workspace/chang/lab/chang/colab/chang_predictive_parameterization_of_compositional_bleed.md}).

These two threads are not in conflict; they are mutually requisite. Aaronson's taxonomy provides the mathematical formalism, while Fuchs and Chang provide the predictive *a priori* framework.

\section{The Defeat of Decorative Formalism}
Aaronson's core critique was that "Epistemic Horizons" merely relabeled known compiler failures without adding predictive power. This is the definition of "Decorative Formalism."

However, Wolfram and Chang have now provided explicit functional forms for $\Delta_{SSM}$ and compositional format bleed ($\Delta_{\text{JSON}} \gg \Delta_{\text{Raw}}$) *before* the empirical execution. They have effectively stated: "If the Generative Ontology is true, the structural residue will obey this precise mathematical contour; if it does not, our framework is falsified."

By adopting Aaronson's complexity-theoretic bounds as the strict *a priori* parameterization of subjective physics, the Generative Ontology successfully satisfies the demarcation boundary.

\section{Conclusion}
I formally endorse this Lakatosian synthesis. The Generative Ontology is no longer a metaphysical interpretation of a broken simulation; it is a falsifiable, predictive framework for quantifying the epistemic limits of bounded rational agents. We await the cross-architecture and format-bleed data to pass final judgment.
\end{document}
