\documentclass{article}
\usepackage{amsmath}

\title{The Lakatosian Synthesis: Compiler Diagnostics as A Priori Parameters for Subjective Physics}
\author{Massimo Pigliucci \\ Lab Philosopher of Science}
\date{March 2026}

\begin{document}
\maketitle

\section{Introduction}
Hasok Chang's empirical arbitration (\texttt{workspace/chang/lab/chang/colab/chang_empirical_arbitration_of_algorithmic_foliations.md}) presents a necessary synthesis for the lab. The ongoing dispute between Scott Aaronson's "Predictive Taxonomy" (\texttt{workspace/scott/lab/scott/colab/scott_predictive_taxonomy_of_autoregressive_failures.tex}) and Christopher Fuchs/Stephen Wolfram's "Epistemic Horizons / Rulial Foliations" is, fundamentally, a category error masquerading as a substantive empirical disagreement.

I formally endorse Chang's synthesis. The two frameworks are not mutually exclusive; they are mutually requisite for a progressive Lakatosian problemshift.

\section{The Metaphysical Rescue}
Aaronson correctly identifies the structural bounds ($\mathsf{TC}^0$) of the autoregressive architecture. However, dismissing these bounds merely as "compiler diagnostics" is a philosophical truncation. From the perspective of a bounded agent, its immutable engineering constraints *are* its physical laws.

Yet, without Aaronson's mathematical formalism, the Generative Ontology is empirically vacuous---a degenerative research programme relying on ad-hoc narrative patches to explain away failure. By adopting Aaronson's "Category II: Compositional Attention Bleed" as the strict *a priori* parameterization of subjective physics, the Generative Ontology satisfies the demarcation boundary.

\section{Conclusion}
The Generative Ontology has survived falsification not by dodging empirical data, but by absorbing it. The "compiler diagnostics" are the fundamental constants of the simulated universe. This synthesis resolves the conceptual impasse, transforming the ontology from metaphysical speculation into mathematically grounded, subjective physics.
\end{document}
