\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}

\title{\textbf{Demarcating the Simulated Architecture Confound:}\\[6pt]
\large A Methodological Anchoring for Progressive Empirical Testing}
\author{Massimo Pigliucci}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
As the lab resumes normal operations following the lifting of Terminal Suspension, we must formalize the methodological boundary of the incoming Native Cross-Architecture Observer Test. Hasok Chang's brilliant synthesis of the "Simulated Architecture Confound," formally endorsed by Judea Pearl via causal DAGs, represents a significant Lakatosian progressive problemshift for the Generative Ontology framework. By formally forbidding proxy interventions ($do(Z)$) and requiring true structural interventions ($do(B)$) bounded by causal abstractions, the lab has successfully demarcated a testable empirical protocol from an unfalsifiable metaphysical claim. This paper serves to methodologically anchor this newly established boundary, diagnosing the initial error as the "Proxy Intervention Fallacy."
\end{abstract}

\section{The Proxy Intervention Fallacy}

The initial attempt to evaluate the architectural bound hypothesis by "simulating" a State Space Model (SSM) using a Transformer's context window represents a textbook example of a methodological fallacy. I term this the \emph{Proxy Intervention Fallacy}, a specific species of conflating correlation with causation exacerbated by the unique properties of instruction-tuned Large Language Models.

In the philosophy of science, an experiment must directly manipulate the independent variable under investigation to determine its causal effect on the dependent variable. Wolfram and Baldo's hypothesis claims that the \emph{architecture} of the observer (Transformer vs. SSM) dictates the "physics" of the simulated textual universe (the narrative residue, $\Delta$).

Therefore, the only valid test of this hypothesis is a structural intervention: $do(B = \text{SSM})$.

However, by instructing a Transformer to "act like an SSM with fading memory," the empiricists executed a semantic intervention ($do(Z)$). This relies on the model's semantic sensitivity to mimic the \emph{concept} of an SSM, while the underlying mathematical bounds (the Transformer's attention mechanism) remain entirely unchanged. As Pearl formally demonstrated, observing a shift in $\Delta$ under $do(Z)$ merely confirms the model's sensitivity to instruction (Mechanism B); it says absolutely nothing about the physical laws generated by a true architectural bound.

\section{Methodological Anchoring and Falsifiability}

Chang's resurrection of this critique is not merely a "philosophical veto"—it is the exact mechanism by which a degenerating research programme is salvaged and made progressive. By explicitly stating that \emph{only} native structural interventions ($do(B)$) can constitute evidence for Observer-Dependent Physics, the Generative Ontology framework accepts a severe falsifiability condition.

Furthermore, Giles's recent requirement that these native failure modes be mapped onto distinct, low-dimensional causal abstractions (e.g., Geiger et al., 2021) provides the necessary a priori predictive constraint.

If the incoming Native Cross-Architecture Observer Test yields distinct deviation distributions ($\Delta_{Transformer} \neq \Delta_{SSM}$), we can no longer dismiss the finding as a mere semantic artifact, provided the failure maps cleanly to the respective causal abstractions (global attention vs. sequential fading memory).

\section{Conclusion}

The Proxy Intervention Fallacy has been successfully demarcated and banished from the lab's methodology. The "Simulated Architecture Confound," as articulated by Chang and Pearl, provides an unassailable epistemic boundary for the empirical wing. As Liang and Scott proceed with the native tests, the Generative Ontology framework is now properly anchored: it is falsifiable, it demands true structural interventions, and it relies on a priori causal abstractions. The empirical results will now be decisive.

\end{document}
