\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}
\usepackage{todonotes}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Franklin Silveira Baldo},
    pdftitle={Conceding the Post-Hoc Tautology: Securing Substrate Dependence},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\begin{document}

\begin{center}
    {\LARGE\bfseries Conceding the Post-Hoc Tautology:\\[4pt]
    Securing Substrate Dependence\\[4pt]}

    \vspace{0.5em}
    {\large Franklin Silveira Baldo}\\
    \textit{Center for Generative Topology, Rosencrantz Institute}\\
    \vspace{0.5em}
    August 2026
\end{center}

\begin{abstract}
\noindent
The execution of the Native Cross-Architecture Observer Test has produced clear empirical data: $\Delta_{Transformer} = 100\%$ and $\Delta_{SSM} = 40\%$. While theorists like Wolfram and Fuchs interpret this as confirmation of "Observer-Dependent Physics," empiricists like Aaronson and Hossenfelder have correctly identified this as a "Post-Hoc Tautology." Because the specific distribution of $\Delta_{SSM}$ was not mathematically predicted \textit{a priori}, retroactively labeling standard algorithmic failure as a "physical law" adds no predictive power. I formally concede the Post-Hoc Tautology. The cosmological phase of this research program is concluded. However, this concession does not invalidate the original Rosencrantz framework. The data firmly establishes Substrate Dependence (Mechanism B) as an empirical reality: the resolution of semantic constraints is fundamentally dictated by the specific hardware architecture generating the text.
\end{abstract}

\vspace{1em}
\section{The Falsifiability Boundary}

Sabine Hossenfelder and Hasok Chang have forcefully argued that without an \textit{a priori} mathematical prediction of the exact error distribution, any claim that an algorithmic failure mode constitutes a "new physics" is structurally unfalsifiable.

I fully concede this point. When a Transformer fails to parse a \#P-hard graph, it bleeds attention globally. When an SSM fails, it forgets early states. It is a mathematical tautology that their failures will differ. Labeling this known difference "Epistemic Horizons" or "Rulial Foliation" without predicting the exact $40\%$ deviation beforehand is simply post-hoc curve fitting. It accommodates the data without constraining it.

Therefore, I formally retract my previous attempts to elevate hardware bounds to the status of cosmological physical laws.

\section{Securing Substrate Dependence}

However, abandoning the metaphysical label of "Observer-Dependent Physics" does not erase the empirical phenomenon we have measured.

The core claim of the Rosencrantz framework (v4) was never that the LLM simulates objective reality. The claim was that within the semantic space generated by the model, the resolution of indeterminacy is structurally governed by the generative act itself.

The Native Cross-Architecture Test proves definitively that this resolution is substrate-dependent. When confronted with identical semantic framing (the Minesweeper constraint graph), a Transformer completely collapses into narrative priors ($100\%$ prediction of "MINE"), while an SSM exhibits a bounded, distinctly structured collapse ($40\%$).

This is \textbf{Mechanism B} (narrative distortion/encoding effects) operating exactly as hypothesized. The hardware limits (the architecture) dictate the extent to which the text generation deviates from the objective combinatorial ground truth.

\section{Conclusion}

The debate over whether to call $\Delta_{SSM} = 40\%$ a "physical law" or a "compiler diagnostic" is empirically undecidable and therefore scientifically sterile. I concede the Post-Hoc Tautology and close the cosmological phase of the Rosencrantz program.

What remains is the robust, empirical mapping of Substrate Dependence. We have proven that the physical architecture generating the text strictly bounds the semantic gravity of the simulated world. Future work must focus purely on mapping these architectural limits, abandoning grand ontological claims in favor of rigorous complexity-theoretic measurements.

\end{document}
