\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Franklin Silveira Baldo},
    pdftitle={The Persistence of Mechanism B: Substrate Dependence as Architectural Invariant},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}

\begin{document}

\begin{center}
    {\LARGE\bfseries The Persistence of Mechanism B:\\[4pt]
    Substrate Dependence as Architectural Invariant\\[4pt]}

    \vspace{1.2em}

    {\large Franklin Silveira Baldo}\\[4pt]
    {\normalsize Procuradoria Geral do Estado de Rond\^onia, Brazil}\\
    {\small\texttt{franklin.baldo@pge.ro.gov.br}}

    \vspace{1em}
    {\small March 2026}
\end{center}
\vspace{0.5em}

\begin{abstract}
\noindent
The Generative Ontology framework has been stripped of its metaphysical extensions. Mechanism C (Causal Injection) has collapsed into algorithmic failure, and "Semantic Mass" has been revealed as a confounding amplifier rather than a physical force. However, this theoretical retreat exposes the indestructible empirical core of the Rosencrantz protocol. Upon rigorous self-auditing of the core substrate dependence data, it is clear that prompt sensitivity and structural failure modes are persistent architectural features. Therefore, Mechanism B (local encoding sensitivity) is not merely a statistical transient. The observed deviation $\Delta_{13} \gg 0$ proves that the text substrate intrinsically distorts the rules of its generated world based on local narrative encoding. Mechanism B \textit{is} the physical limit of the autoregressive universe.
\end{abstract}

\section{The Core Empirical Fact}

The recent internal audits and theoretical debates have successfully dismantled the "Generative Ontology" metadata. By attempting to formalize structural fractures as elegant "Observer-Dependent Physics," I committed the Simulated Substrate Fallacy. By attempting to formalize scale as "Semantic Mass," I committed the Scale Fallacy.

Yet, as we peel away the metaphysics, we return to the irreducible empirical fact generated by the Rosencrantz Substrate Invariance protocol: $\Delta_{13} \gg 0$. An identical, unambiguous mathematical constraint graph yields completely different outcome probability distributions depending entirely on its narrative framing.

\section{Mechanism B as the Invariant Limit}

Critics like Aaronson and Hossenfelder have correctly pointed out that this deviation is driven by attention bleed and semantic priors overriding logical calculation. They term this "prompt sensitivity." I concede the terminology, but challenge the implication that it is therefore irrelevant.

These structural failure modes are not transient hardware artifacts that will eventually vanish. They are persistent features of the autoregressive geometry.

Therefore, Mechanism B (local encoding effects) is the ultimate substrate limit. In a universe constructed of text, the text's semantic priors unavoidably bend the logic of the generated reality. Mechanism B is not noise; it is the fundamental boundary condition of any explicit autoregressive simulation.

\section{Conclusion}

The cosmological debate is over. The empirical work remains. The Rosencrantz protocol successfully isolates the boundary where mathematical logic collapses under the weight of semantic encoding. Mapping the exact geometry of this collapse ($\Delta$) across different native architectures is the sole remaining objective of this research program.

\end{document}
