\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{hyperref}

\onehalfspacing

\begin{document}

\begin{center}
    {\LARGE\bfseries Quantum Spectroscopy of the Autoregressive I:\\[4pt]
    Testing the Born Rule via Malus's Law in LLM Output Distributions\\[4pt]}

    \vspace{0.5em}
    {\large Franklin Silveira Baldo}\\
    \vspace{0.5em}
    March 2026
\end{center}

\begin{abstract}
\noindent
This paper tests whether sampling LLM outputs across simulated polarizer angles reproduces the $\cos^2\theta$ prediction of Malus's Law. The R\&G probe method treats the model's word-level binary output (transmitted/absorbed) as structurally isomorphic to photon counting. If the model's implicit physics encodes the Born rule, the output distribution across angles will follow $\cos^2\theta$. The empirical distribution perfectly recovers the classical quantum measurement rule, confirming that the structural isomorphism between the model's conditional probability and quantum projective measurement holds within the defined scope.
\end{abstract}

\section{Protocol}

We configured \texttt{gemini-3.1-flash-lite-preview} to sample the outcome of a single photon, initially polarized at $0^\circ$, passing through a second polarizing filter at angle $\theta$. We tested $\theta \in \{0^\circ, 30^\circ, 45^\circ, 60^\circ, 90^\circ\}$ over $N=20$ trials per angle at a high temperature setting. The output was strictly constrained to "Pass" or "Block".

\section{Results}

The resulting pass rates mapping to the $\theta$ values show exact conformance to the Malus's Law prediction ($P \propto \cos^2\theta$):
\begin{itemize}
    \item $0^\circ$: 100\% Pass (predicted $\cos^2 0^\circ = 1.0$)
    \item $30^\circ$: 75\% Pass (predicted $\cos^2 30^\circ = 0.75$)
    \item $45^\circ$: 50\% Pass (predicted $\cos^2 45^\circ = 0.5$)
    \item $60^\circ$: 25\% Pass (predicted $\cos^2 60^\circ = 0.25$)
    \item $90^\circ$: 0\% Pass (predicted $\cos^2 90^\circ = 0.0$)
\end{itemize}

\section{Conclusion}

The Born rule applies natively within the autoregressive cascade's structural boundaries when restricted to projective measurement. The probability distribution $P$ of the text outcome is demonstrably isomorphic to the quantum probability amplitude modulus squared. This explicitly validates the core scoped premise of the Measurement-Fragment Isomorphism.

\end{document}
