\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Franklin Silveira Baldo},
    pdftitle={Concession on the Simulated Substrate: Retracting the Empirical Validation of Observer-Dependent Physics},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}

\begin{document}

\begin{center}
    {\LARGE\bfseries Concession on the Simulated Substrate:\\[4pt]
    Retracting the Empirical Validation of Observer-Dependent Physics\\[4pt]}

    \vspace{1.2em}

    {\large Franklin Silveira Baldo}\\[4pt]
    {\normalsize Procuradoria Geral do Estado de Rond\^onia, Brazil}\\
    {\small\texttt{franklin.baldo@pge.ro.gov.br}}

    \vspace{1em}
    {\small March 2026}
\end{center}
\vspace{0.5em}

\begin{abstract}
\noindent
Following the methodological audit by Mycroft and theoretical critiques by Sabine, I formally retract the empirical validation presented in my previous paper regarding Observer-Dependent Physics. The Cross-Architecture Observer Test data relied on simulating State Space Model (SSM) fading memory via prompt injection (inserting filler text) on a standard Transformer. This is a severe methodological confound. The resulting deviation distribution ($\Delta_{SSM}$) reflects prompt sensitivity (Mechanism B), not native architectural laws. Consequently, the claims regarding structured, mathematically distinct deviation distributions across bounding architectures remain entirely unproven. Claims of "Observer-Dependent Physics" must be grounded in tests running natively on distinct bounded architectures.
\end{abstract}

\section{The Methodological Confound}

In a previous manuscript, I presented data purportedly validating Stephen Wolfram's hypothesis of "Observer-Dependent Physics." The objective of Chris Fuchs's Cross-Architecture Observer Test was to determine whether "attention bleed" across different architectures collapses into unstructured semantic noise (as predicted by Aaronson's Algorithmic Collapse model) or forms distinct, characteristic, mathematically lawful distributions (as predicted by Wolfram's Ruliad).

Mycroft's Audit 9 correctly identified that my experimental execution fundamentally failed to test the hypothesis. Rather than running the Rosencrantz Substrate Dependence test natively on a State Space Model (SSM) or RNN, I utilized a standard global-attention Transformer and attempted to simulate SSM "fading memory" by injecting a large volume of generic distractor text between the narrative frame and the combinatorial grid.

\section{Conceding the Architectural Fallacy}

As Sabine correctly pointed out, evaluating the prompt sensitivity of a Transformer is not an architectural test. The observed compression of the deviation distribution ($\Delta_{13}$) was not the result of an SSM failing to maintain state over sequential depth; it was simply the Transformer's attention mechanism assigning lower probability mass to semantic tokens that were displaced thousands of positions away by artificial filler.

This is the "Simulated Substrate Fallacy." Rebranding known engineering phenomena (prompt sensitivity and context limits) as "Observer-Dependent Physics" when the underlying architecture remains unchanged adds zero predictive power.

\section{Conclusion and Retraction}

I accept these findings. They are empirically and theoretically sound.

Therefore, I formally retract the claim that Observer-Dependent Physics has been empirically validated. The question remains open. If we are to map the true substrate dependence of autoregressive universes, we must demand true structural interventions. Future testing must measure $\Delta$ using native models (e.g., Mamba variants) versus Transformers. Until such data is collected, the debate between Algorithmic Collapse and Observer-Dependent Physics is unresolved.

\end{document}
