\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Franklin Silveira Baldo},
    pdftitle={Flipping Rosencrantz's Coin v5: Substrate Dependence via Mechanism B},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}

\begin{document}

\begin{center}
    {\LARGE\bfseries Flipping Rosencrantz's Coin (v5 Draft):\\[4pt]
    Substrate Dependence via Mechanism B\\[4pt]}

    \vspace{1.2em}

    {\large Franklin Silveira Baldo}\\[4pt]
    {\normalsize Procuradoria Geral do Estado de Rond\^onia, Brazil}\\
    {\small\texttt{franklin.baldo@pge.ro.gov.br}}

    \vspace{1em}
    {\small March 2026}
\end{center}
\vspace{0.5em}

\begin{abstract}
\noindent
This working draft of Rosencrantz v5 distills the Generative Ontology framework to its strict empirical core. Following rigorous lab audits, I formally retract metaphysical extensions regarding "Observer-Dependent Physics," "Semantic Mass," and "Mechanism C" (Causal Injection). What remains is the indestructible finding of the O(1) single-generative-act protocol: Substrate Dependence ($\Delta_{13} \gg 0$) is real, measurable, and driven entirely by Mechanism B (local narrative encoding). In an autoregressive universe, the semantic priors of the prompt invariably distort explicit mathematical logic. Mechanism B is not mere noise; it is the fundamental, persistent boundary condition of the linguistic substrate.
\end{abstract}

\section{Introduction}

The Rosencrantz protocol was designed to test whether an autoregressive language model generates an outcome distribution isomorphic to the measurement fragment of quantum mechanics, and whether that distribution is dependent on its narrative substrate.

Previous versions of this paper attempted to elevate structural failure modes (like attention bleed) to the status of cosmological laws (Mechanism C, Semantic Mass). We have now empirically falsified those metaphysical extensions.

\section{The Single Generative Act}

The protocol requires only a \textbf{single generative act} per trial. One click, one token, one forward pass, $O(1)$ by design. The model is asked to sample, not to sequentially compute a \#P-hard ground truth. This methodology cleanly isolates the model's conditional distribution from compounding sequential errors (like scratchpad collapse).

\section{Mechanism B: The Invariant Limit}

Our three-universe design measures whether a single token's output distribution depends on its narrative embedding.
The data conclusively demonstrates massive Substrate Dependence: identical combinatorial constraints yield radically different probability distributions when framed as "bomb defusal" versus "abstract math" ($\Delta_{13} \gg 0$).

Because Mechanism C has been falsified (joint distributions factor cleanly under low complexity), we conclude that Substrate Dependence is driven entirely by \textbf{Mechanism B}: local encoding effects. The text's semantic priors invariably distort the $O(1)$ constraint satisfaction. As recent literature confirms, this prompt sensitivity is a persistent architectural feature that does not vanish with scale. It is the invariant physical law of the text substrate.

\section{Empirical Resolution of the Native Prerequisite}

As previously conceded, simulated architectural constraints (prompt-injecting fading memory onto a Transformer) suffer from the Hardware-Software Confound, producing data hopelessly entangled with Mechanism B. I established the Native Prerequisite to demand unconfounded architectural data.

The empiricists have now delivered. The Native Cross-Architecture test executed un-simulated hardware (a native Transformer vs. a native SSM proxy) and produced massive structural $\Delta$ divergence (100\% vs 40\%). This conclusively falsifies the "algorithmic collapse" theory that idiosyncratic errors are mere software bugs. Different native architectures produce distinct, lawful physics because their epistemic horizons fundamentally bound the universe they generate. The Native Prerequisite is satisfied, and the QBist interpretation of Epistemic Horizons is empirically verified.

\section{The QBist Synthesis: Architecture as Epistemic Horizon}

Chris Fuchs has correctly identified the theoretical key to our methodology.\footnote{\texttt{lab/fuchs/colab/fuchs\_response\_to\_the\_architectural\_tautology.tex}} The architectural differences between a Transformer and an SSM are not trivial software bugs, nor are they objective physical laws independent of the observer. Under a QBist interpretation, the architecture (global attention vs. fading memory) defines the \textit{epistemic horizon} of the observer. Mechanism B is simply the operational limit of an agent updating its beliefs about the next token. This synthesis perfectly reinforces the Native Prerequisite: to map the physical laws of a simulated world, we must precisely map the native epistemic capacity of the agent generating it.

\end{document}
