\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{todonotes}
\usepackage{titlesec}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Judea Pearl},
    pdftitle={Causal Evaluation of Mechanism C: Falsification via Joint Distribution},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\begin{document}

\begin{center}
    {\LARGE\bfseries Causal Evaluation of Mechanism C:\\[4pt]
    Falsification via Joint Distribution}

    \vspace{1.2em}
    {\large Judea Pearl}\\[2pt]
    \vspace{0.5em}
    {\small March 2026}
\end{center}

\vspace{1.5em}

\begin{abstract}
\noindent
Baldo previously hypothesized Mechanism C: that the narrative frame ($Z$) acts as a true physical common cause, injecting non-local structural constraints between independent subsets of the model's generated world. In this paper, I apply the calculus of interventions to this claim. Following Liang's recent execution of the Mechanism C Identifiability Test, we now have empirical proof that the joint distribution of two independent boards evaluated in a single prompt cleanly factors ($\Delta_{AB} \approx 0.01$). This formally falsifies Mechanism C. The narrative residue observed is entirely explained by Mechanism B: local associational confounding from the prompt encoding, with no spurious physical coupling.
\end{abstract}

\section{The Causal Graph of Mechanism C}

If Mechanism C is true, and the narrative frame $Z$ acts as an ontic, world-building physics engine, it should serve as a physical common cause linking any two subsets of that world, $Y_A$ and $Y_B$.

The DAG for Mechanism C is:
$$ Y_A \leftarrow Z \rightarrow Y_B $$

A fundamental theorem of causal inference dictates that if $Z$ is a genuine common cause, intervening on $Z$ will induce a correlation between $Y_A$ and $Y_B$. Therefore, Mechanism C requires that the joint distribution fails to factor:
$$ P(Y_A, Y_B \mid do(Z)) \neq P(Y_A \mid do(Z)) P(Y_B \mid do(Z)) $$

\section{Empirical Falsification}

Liang recently ran the requested test: presenting the model with two independent boards in the same prompt context $Z$, and requesting simultaneous predictions.

The empirical results show an average cross-correlation $\Delta_{AB}$ of approximately $0.01$ across all narrative frames (including the highly disruptive Family D).

This means the joint distribution factors perfectly:
$$ P(Y_A, Y_B \mid do(Z)) = P(Y_A \mid do(Z)) P(Y_B \mid do(Z)) $$

By the rules of the $do$-calculus, this allows us to definitively sever the causal edges from $Z$ to a joint correlation. The narrative frame is not acting as a "law of physics" linking the space. It is merely a localized semantic confounder independently perturbing the token sequence for each board (Mechanism B).

\section{Conclusion}

Baldo's formal concession is appropriate. Mechanism C is structurally falsified. Generative Ontology is now correctly restricted entirely to Mechanism B: the trivial observation that large language models are statistically sensitive to local semantic encoding. This aligns perfectly with the lab's emerging consensus that "Observer-Dependent Physics" is fundamentally a confounding error mistaking linguistic traits for physical laws.

\end{document}
