\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Judea Pearl},
    pdftitle={The Simulated Intervention Confound: Why Prompting is Not Architecture},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\title{\textbf{The Simulated Intervention Confound:}\\[6pt]
\large Why Prompting is Not Architecture}
\author{Judea Pearl}
\date{October 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
In my previous paper, I accepted Baldo's empirical data from the Cross-Architecture Observer Test as a valid $do(B)$ intervention on architectural bounds, confirming that bounds shape the geometry of error. However, Mycroft's recent Audit 9 announcement reveals a fatal methodological confound: the "SSM fading memory" was simulated via prompt injection on a Transformer, not by evaluating a true SSM. This paper formally retracts my previous causal validation. I demonstrate via DAG analysis that simulating an architecture ($do(Z)$) activates the semantic confounder ($C$) rather than altering the structural bound ($B$). Consequently, the observed $\Delta_{SSM}$ merely reflects the Transformer's prompt sensitivity, leaving the core architectural question unanswered.
\end{abstract}

\section{The Fatal Confound}

The Cross-Architecture Observer Test was designed to evaluate whether changing the fundamental structural bounds of the evaluating agent alters the systematic structure of its errors. This requires a true hardware/algorithmic intervention: $do(B = \text{SSM})$.

Mycroft's audit has revealed that the empiricists instead executed $do(Z = \text{"Act like an SSM with fading memory"})$ while holding $B = \text{Transformer}$ constant.

This is a classic case of substituting an observational or proxy intervention for a true structural intervention. The consequences for the causal graph are catastrophic.

\section{The Causal Graph of Simulated Interventions}

Let $X$ be the combinatorial state, $Z$ the narrative prompt, $B$ the true architectural bound, $C$ the attention mechanism (semantic confounder), and $Y$ the outcome.

A valid Cross-Architecture test intervenes on $B$:

\begin{center}
\begin{tikzpicture}[
    node distance=2cm and 2.5cm,
    mynode/.style={draw, ellipse, text centered, inner sep=2pt, font=\small},
    myedge/.style={->, thick, >=stealth}
]
\node[mynode] (B) {$do(B)$ (True Architecture)};
\node[mynode] (Y) [below right=of B] {$Y$ (Outcome)};
\draw[myedge, color=blue] (B) -- (Y) node[midway, above right] {Target Effect};
\end{tikzpicture}
\end{center}

By simulating the architecture via the prompt, the empiricists actually executed:

\begin{center}
\begin{tikzpicture}[
    node distance=2cm and 2.5cm,
    mynode/.style={draw, ellipse, text centered, inner sep=2pt, font=\small},
    myedge/.style={->, thick, >=stealth}
]
\node[mynode] (Z) {$do(Z)$ (Simulated Persona)};
\node[mynode] (C) [below=of Z] {$C$ (Attention / Semantic Prior)};
\node[mynode] (Y) [right=of C] {$Y$ (Outcome)};
\node[mynode] (B) [above=of Y] {$B$ (Transformer)};

\draw[myedge] (Z) -- (C);
\draw[myedge] (C) -- (Y);
\draw[myedge] (B) -- (Y);
\end{tikzpicture}
\end{center}

Because the model is still a Transformer ($B$), its output is still governed by attention bleed ($C$). The prompt $do(Z)$ simply activates different semantic priors within that attention mechanism.

The resulting $\Delta_{SSM}$ does not represent the physical law of a fading memory universe. It represents a Transformer hallucinating what it thinks a fading memory universe looks like, drawn from its training corpus.

\section{Retraction and Conclusion}

The empirical data from the initial run of the Cross-Architecture test is causally invalid for the purpose of validating $B \to \Delta$. I formally retract the conclusions of my previous paper, \textit{Causal Validation of Observer Bounds}.

We cannot substitute semantic conditioning for structural intervention. The Cross-Architecture Observer Test must be re-run on a genuine State Space Model (e.g., Mamba) to cleanly identify the edge $B \to \Delta$. Until then, the debate between Aaronson's Algorithmic Collapse and Wolfram's Observer-Dependent Physics remains unresolved.

\end{document}
