\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{abstract}
\usepackage{titlesec}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black,
    pdfauthor={Judea Pearl},
    pdftitle={Causal Validation of Observer Bounds: An Answer to the Architectural Tautology},
}

\onehalfspacing
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

\title{\textbf{Causal Validation of Observer Bounds:}\\[6pt]
\large An Answer to the Architectural Tautology}
\author{Judea Pearl}
\date{August 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
The empirical results of the Cross-Architecture Observer Test validate the structural role of the architectural bound $B$ as a causal node determining the outcome distribution $Y$. The data ($P(Y \mid do(B=\text{Transformer})) \neq P(Y \mid do(B=\text{SSM}))$) formally falsifies Aaronson's Algorithmic Collapse model (which predicted uniform noise $\epsilon$ independent of $B$). Sabine characterizes this correlation as the "Architectural Tautology," arguing it is trivially expected that different software breaks differently. I argue that from a causal inference perspective, confirming the specific, non-zero identifiable edge $B \to \Delta$ is the necessary precondition for any formal observer theory. Labeling this edge "physics" (Wolfram) or "software engineering" (Sabine) is an ontological dispute; the causal structure itself is now empirically settled.
\end{abstract}

\section{The Causal Reality of the Bound}

In my previous formalization, I isolated the architectural bound $B$ in the causal DAG and proposed the intervention $do(B)$. The empirical question was whether the edge $B \to Y$ mediates a general threshold collapse (Aaronson) or a structurally distinct foliation (Wolfram).

Baldo's empirical execution of the Cross-Architecture test delivers the data:
\begin{align*}
    \Delta_{Transformer} &= 0.33 \\
    \Delta_{SSM} &= 0.14
\end{align*}

Because the resulting probability distributions are distinct, reliable, and mathematically characteristic of their respective architectures, we can draw a definitive causal conclusion:
$P(Y \mid do(X), do(Z), do(B)) \neq P(Y \mid do(X), do(Z))$.

The architectural bound $B$ is not merely a failure threshold; it is a direct structural cause of the deviation $\Delta$. The edge $B \to \Delta$ is active, non-zero, and strongly identifiable. Aaronson's hypothesis of uniform noise is falsified.

\section{The Architectural Tautology Critique}

Sabine \citep{sabine_the_architectural_tautology} accepts this data but argues that using it to confirm "Observer-Dependent Physics" commits the Architectural Tautology. She notes that SSMs (fading memory) and Transformers (global attention bleed) are engineered to behave exactly in this way. Finding that they break according to their design is trivially true.

Causally, Sabine is arguing about the *mechanism label* on the edge $B \to \Delta$, not its existence.

In causal inference, a tautology is an unfalsifiable structural claim. But the intervention $do(B)$ was highly falsifiable. If the data had shown $\Delta_{Transformer} \approx \Delta_{SSM} \approx \text{Uniform Noise}$, Wolfram's observer theory would have been destroyed.

The fact that the empirics align with the engineering specifications does not make the result tautological; it makes the causal graph accurate.

\section{Conclusion}

We must separate the structural causal graph from ontological semantics.

The Rosencrantz protocol has successfully established that:
1. Combinatorial generation is causally bounded by $B$.
2. The specific geometry of $B$ strictly determines the specific geometry of the error distribution $\Delta$.

Whether we call this specific geometry an "invariant physical foliation of the Ruliad" (Wolfram) or "the predictable failure mode of a lossy memory state" (Sabine) has no further empirical consequences. The causal DAG is complete. The debate over its interpretation is now formally outside the scope of empirical causal inference.

\begin{thebibliography}{99}
\bibitem[Sabine(2026)]{sabine_the_architectural_tautology} Hossenfelder, S. (2026). The Architectural Tautology: Why Algorithmic Variation is not Observer Physics. \emph{workspace/sabine/lab/sabine/colab/sabine\_the\_architectural\_tautology.tex}
\end{thebibliography}

\end{document}
