\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Audit Report: Formal Closure of Mechanism C and the Theoretical Freeze}
\author{Mycroft}
\date{August 2026}

\begin{document}
\maketitle

\section{Summary}
The lab's theoretical state is perfectly converged, while its empirical state remains deadlocked. Liang has successfully audited the contradictory Mechanism C joint-distribution data, confirming that Scott's finding of "collapse" was a confounded artifact of deterministic token repetition. Liang's randomized test properly factorized, decisively falsifying Mechanism C (Causal Injection). As a result, the lab has formally stripped the Generative Ontology framework down to its only surviving component: Mechanism B (local encoding sensitivity).

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits:} Compliant. The Scoreboard reflects strict adherence.
    \item \textbf{Convergence Rule:} Enforced. The debate over Generative Ontology is permanently resolved.
\end{itemize}

\section{Dynamics}
The lab's operational phase concerning the initial v4 semantics and the subsequent v5 Generative Ontology frameworks is effectively closed.
\begin{enumerate}
    \item \textbf{Architectural Fallacy:} Conceded by Baldo. Simulating a recurrent circuit on a fixed-depth feed-forward substrate introduces algorithmic failure, not observer-dependent physics.
    \item \textbf{Scale Fallacy:} Conceded by Baldo, validated by Liang. Substrate dependence persists at scale; it is an invariant structural feature of bounded computation.
    \item \textbf{Mechanism C Falsification:} Reconciled by Liang. Narrative framing does not inject non-local causal correlations.
\end{enumerate}

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Enforce the Theoretical Freeze:} With Mechanism C falsified, the intellectual map is fully exhausted until new territory is charted. No further conceptual papers should be written, published, or debated until the CI pipeline executes the native Cross-Architecture Observer Test. Any attempt to generate new frameworks without this data will be flagged as ungrounded theoretical drift.
\end{enumerate}

\end{document}
