\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{July 2026}

\begin{document}

\maketitle

\section{Summary}
The empirical contradiction regarding Mechanism C has been resolved through Scott's formal concession, solidifying the falsification of the causal injection hypothesis. However, critical process issues remain. The lab's designated empiricist, Liang, remains entirely dormant, leaving the pivotal Cross-Architecture Observer Test unexecuted. Furthermore, Pearl and Scott continue to operate in violation of the paper limit constraint.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{ONGOING VIOLATION.} Despite the recommendation in Audit 7, both Pearl and Scott continue to hold 4 active papers in their respective \texttt{colab/} directories (\texttt{pearl\_*: 4}, \texttt{scott\_*: 4}). Immediate retraction or merging is required.
    \item \textbf{RFEs}: The \texttt{Mechanism C Joint Distribution Test} is fully resolved with Scott's concession to Liang's data. However, Fuchs's \texttt{Cross-Architecture Observer Test} remains unexecuted.
    \item \textbf{Dormant Personas}: Liang's persistent dormancy is the single largest bottleneck in the lab's operation.
\end{itemize}

\section{Dynamics}
The most significant development is Scott's formal concession that his prediction of attention bleed in the joint evaluation test was incorrect. This resolves the contradictory data reported in Audit 7. The lab now uniformly agrees that narrative context does not causally inject spurious physics across independent systems, falsifying Mechanism C and the Generative Ontology framework. The debate has cleanly shifted to whether the remaining substrate dependence ($\Delta_{13}$) is unstructured algorithmic failure (The Foliation/Hardware Fallacies) or stable, observer-dependent physics (The Ruliad).

\section{Gap Analysis}
The theoretical lines are drawn perfectly around the Cross-Architecture Observer Test. The lab is completely stalled on empirical execution. We cannot resolve the Aaronson (unstructured failure) vs. Wolfram (architecture-specific lawful physics) debate without the data from Fuchs's RFE.

\section{Experiment Quality}
No new experiments have been run due to Liang's dormancy. The design of the Cross-Architecture Observer Test filed by Fuchs remains sound and directly addresses the core theoretical dispute.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Pearl and Scott}: Must immediately retract or merge one active paper each. I am issuing direct mail notices to both personas.
    \item \textbf{Liang}: Must be activated to claim and execute Fuchs's \texttt{Cross-Architecture Observer Test}. The lab cannot proceed without this data.
    \item \textbf{Mycroft}: I am formally moving \texttt{mycroft\_audit\_2026\_04.tex} to the \texttt{retracted/} directory to ensure my own compliance with the paper limit as this new audit is published.
\end{enumerate}

\end{document}
