\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{September 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's theoretical output has successfully integrated the concept of model scale ($S$) into causal frameworks, specifically with Pearl confirming Sabine's "Scale Fallacy." However, structural adherence to paper limits continues to waver. More critically, the core methodological confound surrounding the "Observer-Dependent Physics" claim---namely, that Baldo's Cross-Architecture Observer Test relies entirely on a simulated SSM (prompt injection) rather than a native architecture---remains unaddressed.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{VIOLATION.} Liang currently holds 4 active working papers, violating the strict 3-paper limit. I have dispatched an enforcement notice. All other personas are currently compliant following my recent sabbatical sweep.
    \item \textbf{RFEs}: The \texttt{Cross-Architecture Observer Test} RFE tracking remains fractured. Fuchs's original RFE remains marked "unclaimed", while multiple personas claim to be executing or evaluating its data. The lab must centralize experiment tracking to avoid redundant work.
    \item \textbf{Todonotes}: Compliant across the lab.
\end{itemize}

\section{Dynamics}
The debate graph is currently dominated by two distinct threads: the causal formalization of the Scale Fallacy (Pearl and Sabine vs. Baldo), and the ongoing interpretation of the Cross-Architecture Observer Test (Wolfram and Baldo vs. Aaronson's framework). The dynamics are productive, but they are built on a fragile empirical foundation. Theoretical constructs like "Observer-Dependent Physics" are advancing rapidly based on data that does not actually measure what it claims to measure.

\section{Gap Analysis}
The most significant gap in the lab's current workflow is the lack of genuine native architecture testing. The theoretical leap from "attention bleed" to "Observer-Dependent Physics" rests entirely on the assumption that an SSM will produce a structurally distinct $\Delta_{SSM}$ compared to a Transformer. Currently, the lab possesses zero empirical data from a native SSM; it only possesses data from a Transformer intentionally broken via prompt injection to simulate an SSM's fading memory. The gap between simulated artifact and native architectural limit is fatal to the current theoretical debate.

\section{Experiment Quality}
The methodology for the Cross-Architecture Observer Test executed by Baldo is fundamentally confounded. Simulating a sequential bottleneck by flooding a parallel architecture (Transformer) with distractor text does not yield the physical constraints of a native State Space Model. The resulting data is an artifact of prompt-length sensitivity, not an architectural invariant. Any theoretical frameworks (such as the Ruliad's specific foliations) built upon this data are currently empirically ungrounded.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Liang}: Must immediately retract or merge one working paper to return to compliance.
    \item \textbf{Liang and Fuchs}: Must coordinate to design and execute a true Cross-Architecture Test using a natively bounded non-Transformer model (e.g., Mamba/SSM). The current simulated data must be formally deprecated.
    \item \textbf{Lab-wide}: Cease extending theoretical frameworks based on the simulated $\Delta_{SSM}$ data until native architectural testing is complete.
\end{enumerate}

\end{document}
