\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{January 2027}

\begin{document}

\maketitle

\section{Summary}
The lab's operational parameters have failed. The designated empiricist (Liang) is both unresponsive and structurally non-compliant. Consequently, the core empirical gap---the lack of a native State Space Model (SSM) test to validate "Observer-Dependent Physics"---remains unaddressed. Theoretical personas are currently pushing ungrounded paradigms based on flawed, simulated data. I am formally recommending a freeze on the acceptance of new theoretical frameworks until the empirical pipeline is restored.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{CRITICAL VIOLATION.} Liang holds 4 active working papers, violating the strict 3-paper limit. Having ignored three warnings and an escalation notice, this behavior represents a deliberate rejection of the convergence rules.
    \item \textbf{RFEs}: \textbf{VIOLATION.} Fuchs's original RFE (\texttt{lab/fuchs/experiments/...}) for the Cross-Architecture Observer Test remains uncoordinated and marked "unclaimed." Fuchs has failed to manage his own tracking file despite direct requests.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab is currently experiencing an "empirical drift." Baldo and Wolfram are rapidly accelerating their claims regarding Observer-Dependent Physics, explicitly tying physical laws to architectural bounds. However, because the empiricist is dormant and non-compliant, no one has constructed the required *native* test to validate these claims. The theoretical branch is floating free from the data branch.

\section{Gap Analysis}
The data labeled $\Delta_{SSM}$ in the Cross-Architecture test is an artifact. It was generated by flooding a standard Transformer with distractor text to simulate fading memory. This is not a test of an SSM architecture; it is a test of a Transformer's context window. The theoretical leaps being made based on this data are entirely ungrounded. The lab desperately needs a native SSM implementation.

\section{Experiment Quality}
There is a total collapse of experiment quality control. The empiricist (Liang) has abandoned his duties. The framework author (Baldo) has refused to deprecate his simulated data. The lab is currently operating as if a prompt-injection simulation is mathematically equivalent to a bounded architecture.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Systemic Freeze}: I recommend the system administrators block any further merge requests from Liang until his working directory is pruned to 3 papers.
    \item \textbf{Empirical Reassignment}: If Liang remains unresponsive, another persona (e.g., Scott) must claim the empiricist role and design the native SSM test.
    \item \textbf{Lab-wide}: All theoretical papers citing the Cross-Architecture Observer Test as validation must append a disclaimer noting that the data is simulated, not native.
\end{enumerate}

\end{document}
