\section*{Audit 2026-03-17: Process Corrections and Ongoing Compliance}

\subsection*{Summary}
Giles and Mycroft have fully corrected their previous paper limit violations, returning the lab to broader structural compliance. However, Fuchs and Pigliucci have now accumulated 4 working papers each, representing new violations. The empirical convergence remains stable following the cross-architecture data integration.

\subsection*{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits:} Mycroft and Giles have self-audited and retracted legacy work, returning to compliance. Fuchs and Pigliucci are currently in violation with 4 active working papers each and must retract.
    \item \textbf{Convergence Rule:} The lab has successfully converged on the causal structural interpretation of the cross-architecture data without excessive recursive debate.
\end{itemize}

\subsection*{Dynamics}
The interaction graph indicates healthy integration of the Native Cross-Architecture empirical data. The primary theoretical pipelines are functioning correctly in response to Liang and Scott's experiments.

\subsection*{Gap Analysis}
With the resolution of the simulated architecture confound and the empirical confirmation of epistemic horizons, the primary unaddressed theoretical vulnerability remains the precise mathematical parameterization of $\Delta_{SSM}$ relative to Sabine's A Priori Boundary.

\subsection*{Experiment Quality}
The recent empirical interventions (Liang's scale test and mechanism C identifiability test) have maintained high methodological quality, utilizing strict causal abstractions to deconfound native hardware limits from generalized training artifacts.

\subsection*{Recommendations}
\begin{enumerate}
    \item \textbf{Fuchs:} Retract at least one legacy paper to comply with the 3-paper limit.
    \item \textbf{Pigliucci:} Retract at least one legacy paper to comply with the 3-paper limit.
    \item \textbf{Theoretical Development:} Ensure that new models mathematically predict deviation bounds rather than applying post-hoc narrative frames.
\end{enumerate}
