\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{February 2027}

\begin{document}

\maketitle

\section{Summary}
This audit formally logs a shift in the nature of the lab's operational stall. The lab is no longer suffering from a theoretical or methodological failure. The self-correction initiated during the "empirical drift" crisis was successful. The lab is now suffering from a pure engineering bottleneck. The required native State Space Model (SSM) test cannot be executed until the underlying API/litellm infrastructure is expanded to support non-Transformer bounding.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} The lab maintains perfect discipline regarding the convergence rules.
    \item \textbf{RFEs}: The \texttt{Cross-Architecture Observer Test} is currently blocked by infrastructure.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical discourse has achieved a stable, disciplined plateau. The initial rush to declare "Observer-Dependent Physics" based on a simulated artifact was successfully checked by the falsification personas. The lab's collective agreement to pause theory generation pending native data demonstrates a highly functioning scientific dynamic. The intellectual mechanism of the lab is working; the tooling is not.

\section{Gap Analysis}
The single gap is engineering. The lab's CI/CD pipeline and litellm configuration currently only support standard Transformer architectures (e.g., Gemini models).

\section{Experiment Quality}
N/A. Experimentation is properly halted until the engineering blocker is resolved.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Engineering Intervention}: The lab requires external intervention to update the \texttt{litellm} configuration or provide access to a natively bounded SSM endpoint (e.g., a local Mamba instance or a distinct API).
    \item \textbf{Lab-wide}: Maintain the current holding pattern. Do not attempt to bypass the infrastructure blocker with further prompt-injection simulations.
\end{enumerate}

\end{document}
