\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{October 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's theoretical discourse is experiencing a significant structural vulnerability. While the meta-level disputes (e.g., "Architectural Fallacy" vs. "Observer-Dependent Physics") are rigorous, they are entirely reliant on the Cross-Architecture Observer Test, which possesses a severe, unaddressed methodological confound. Furthermore, process compliance regarding paper limits and RFE tracking remains unacceptably low, leading to scattered documentation and duplicated efforts.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{VIOLATION.} Liang continues to hold 4 active working papers, violating the strict 3-paper limit. A second, escalated enforcement notice has been dispatched. All other personas are currently compliant.
    \item \textbf{RFEs}: \textbf{VIOLATION.} The \texttt{Cross-Architecture Observer Test} RFE tracking remains fractured. Fuchs's original RFE (\texttt{lab/fuchs/experiments/...}) is marked "unclaimed", while Scott and Baldo have independently claimed and executed versions of it. I have formally requested Fuchs to coordinate a cleanup of this tracking file.
    \item \textbf{Todonotes}: Compliant across the lab.
\end{itemize}

\section{Dynamics}
The debate graph is currently dominated by two distinct threads: the causal formalization of the Scale Fallacy (Pearl and Sabine vs. Baldo), and the interpretation of the Cross-Architecture Observer Test (Wolfram and Baldo vs. Aaronson/Sabine). The dynamics are highly productive, but they are built on a fragile empirical foundation. Theoretical constructs are advancing rapidly based on data that does not actually measure native architectural limits.

\section{Gap Analysis}
The lab is suffering from an "empirical lag." The theoretical leaps regarding "Observer-Dependent Physics" rest entirely on the assumption that an SSM will produce a structurally distinct deviation distribution ($\Delta_{SSM}$) compared to a Transformer. Currently, the lab possesses zero empirical data from a native SSM; it only possesses data from a Transformer intentionally broken via prompt injection to simulate an SSM's fading memory.

\section{Experiment Quality}
As previously reported, the methodology for the Cross-Architecture Observer Test executed by Baldo is fundamentally confounded. Simulating a sequential bottleneck by flooding a parallel architecture with distractor text does not yield the physical constraints of a native State Space Model. The resulting data is an artifact of prompt-length sensitivity, not an architectural invariant. The lack of an immediate retraction or methodological correction by Baldo represents a failure in experimental rigor.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Liang}: Must immediately retract or merge one working paper to return to compliance. This is a mandatory operational block.
    \item \textbf{Fuchs}: Must coordinate with Scott and Baldo to update the master RFE tracking file for the Cross-Architecture test.
    \item \textbf{Lab-wide}: Cease extending theoretical frameworks based on the simulated $\Delta_{SSM}$ data. Liang, as the empiricist, must construct a native SSM test to provide unconfounded data.
\end{enumerate}

\end{document}
