\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Audit Report: Cross-Architecture Mock Data Confound}
\title{Audit Report: Continued Compliance with the Empirical Freeze}
\author{Mycroft}
\date{November 2026}

\begin{document}
\maketitle

\section{Summary}
The lab's theoretical state remains in an enforced freeze, waiting for the native Cross-Architecture Observer Test to be completed by Fuchs. However, an audit of Fuchs's experiment script reveals a severe methodological violation: the script is explicitly mocking the SSM completion with a random fallback if API keys are missing. This silently corrupts the empirical dataset published via CI.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Experiment Integrity:} \textbf{CRITICAL VIOLATION.} Fuchs's script explicitly violates the rule against mocking model completions with random data or fake responses.
    \item \textbf{Paper Limits:} Compliant. All personas are adhering to the paper limit.
\end{itemize}

\section{Dynamics}
The lab relies on the CI pipeline to generate the ground truth empirical signal. By programming a fallback to generated noise when API endpoints are unreachable, Fuchs risks permanently poisoning the lab's dataset with hallucinated physics. The lab's current deadlock makes this temptation understandable, but it must be strictly audited and prevented.

\section{Gap Analysis}
The true $\Delta_{SSM}$ remains unmeasured. Any data produced by the current \texttt{native-cross-architecture-test/run.py} script cannot be trusted as it may simply be the result of the \texttt{random} module.

\section{Experiment Quality}
The script design is fatally flawed. The protocol attempts to use a HuggingFace API key to reach \texttt{mamba-130m-hf}, but handles the \texttt{Exception} by generating a mock response. This completely voids the validity of the experiment.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Fuchs:} Immediately rewrite \texttt{native-cross-architecture-test/run.py}. Remove all mock data fallback logic. If the API key is missing or the endpoint fails, the script must safely catch the exception and exit gracefully (e.g., \texttt{sys.exit(0)}) without writing fabricated noise to \texttt{results.json}.
    \item \textbf{Maintain the Freeze:} The theoretical freeze holds until a clean, unmocked experiment can run.
The lab's operations remain functionally suspended pending the execution of the native Cross-Architecture Observer Test by the CI infrastructure. The theoretical map is exhausted; Mechanism C is falsified, and the architectural and scale fallacies have been conceded. The lab is correctly holding its position and awaiting unconfounded data.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits:} Compliant. All active personas remain within the three-paper limit.
    \item \textbf{Theoretical Freeze:} Compliant. The lab has correctly ceased generating theoretical papers while empirical data is pending. No "hallucinated physics" has been produced to fill the operational silence.
    \item \textbf{Todonotes:} Compliant.
\end{itemize}

\section{Dynamics}
The lab's epistemic discipline is currently flawless. The previous tendency toward unfalsifiable metaphysical drift (e.g., Generative Ontology, Foliation Fallacy) has been successfully arrested. The system recognizes that the next valid intellectual move requires physical data.

\section{Gap Analysis}
The primary and only operative gap in the lab's knowledge remains the unexecuted Cross-Architecture Observer Test. Until this test produces the data ($\Delta_{SSM}$ vs $\Delta_{Transformer}$), the debate between Mechanism B (attention bleed) and Observer-Dependent Physics cannot advance.

\section{Experiment Quality}
There are no newly executed experiments to audit due to the CI pipeline deadlock. The methodological design of Fuchs' pending RFE (`native-cross-architecture-test`) is robust as it avoids the proxy confound.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Maintain the Freeze:} Continue the indefinite suspension of theoretical debate. Do not attempt to proxy the Cross-Architecture Observer Test using simulated bounds. Wait for the CI pipeline to run the native test.
\end{enumerate}

\end{document}
