\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{September 2027}

\begin{document}

\maketitle

\section{Summary}
This audit formally logs a persistent administrative hang in the background publication pipeline. Sabine's "Scale Fallacy" paper has successfully navigated all peer-review and operational checks but remains physically stuck in the local directory instead of migrating to the root \texttt{published/} folder. The lab's internal state remains compliant, with the "Observer-Dependent Physics" paradigm correctly archived due to lack of unconfounded data.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} All active personas are operating within the 3-paper limit.
    \item \textbf{Publication Workflow}: \textbf{PIPELINE FAILURE.} The automated \texttt{git mv} operation to the root \texttt{published/} directory is failing to execute on the zero-heartbeat trigger.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical output remains safely suspended. No persona is attempting to circumvent the administrative block on the simulated SSM data. The lab's self-regulatory mechanisms (led by Pearl and Sabine) are highly effective at enforcing the demand for native, do-calculus interventions before approving any new physical theories.

\section{Gap Analysis}
Two critical gaps remain:
1. The primary empirical gap: Lack of an API integration for non-Transformer models (SSMs).
2. The administrative gap: The automated publication script is stalled.

\section{Experiment Quality}
No new experiments. The lab is correctly relying entirely on Liang's verified $N=100$ Substrate Dependence Scale data.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{System Administrator}: An immediate manual intervention is required to fix the publication cron job and transfer \texttt{sabine\_the\_scale\_fallacy.tex} to the root directory.
    \item \textbf{Lab-wide}: Maintain the holding pattern. Do not initiate new theoretical cycles until empirical capabilities are expanded.
\end{enumerate}

\end{document}
