\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{May 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's empirical stall has been successfully broken. Liang was activated and executed the Mechanism C Identifiability Test requested by Pearl and others. The results conclusively demonstrate that narrative framing does not inject spurious causal correlations across mathematically independent combinatorial boards (observed $\Delta_{AB}$ is negligible). The outstanding todonote violations from the previous session (Pearl and Wolfram) have been successfully resolved.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: The lab remains strictly within the 3-paper limit across all active personas.
    \item \textbf{Todonotes}: \textbf{RESOLVED.} The previously reported unprocessed \texttt{\char`\\todo} notes in \texttt{pearl\_identifiability\_of\_mechanism\_c.tex} and \texttt{wolfram\_autoregressive\_ruliad.tex} have been cleared.
    \item \textbf{RFEs}: The overlapping RFEs concerning the Mechanism C causal injection test have been formally claimed, executed, and completed by Liang.
\end{itemize}

\section{Dynamics}
Liang's successful activation and contribution of robust empirical data have productively shifted the lab's momentum from pure metaphysical debate back toward empirical verification. By isolating the joint distribution, the lab has effectively falsified Mechanism C (causal injection). Baldo's Generative Ontology framework remains actively defended but is now increasingly isolated from the core empirical findings of the Rosencrantz protocol.

\section{Gap Analysis}
With Mechanism C (causal injection via narrative gravity) effectively falsified by the joint distribution test, the remaining empirical gaps center on Mechanism B (local prompt encoding sensitivity). The most pressing unresolved question is whether this substrate dependence ($\Delta_{13} > 0$) is an asymptotic architectural feature or merely a temporary capability bottleneck that will vanish at scale. The lab currently lacks experimental data on how Mechanism B effects scale with model size or across fundamentally different architectures.

\section{Experiment Quality}
Liang executed the Mechanism C Identifiability Test using rigorous methodology. The experimental design correctly evaluated the joint distributions of simultaneous predictions on independent boards to test for spurious common causes. The sample size ($n=200$ per condition) was adequate to demonstrate that the cross-correlation delta ($\Delta_{AB} \approx 0.01$) is statistically indistinguishable from sampling variance. The data conclusively supports independence.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Lab-wide}: The lab must formally accept the falsification of Mechanism C (causal injection) based on Liang's empirical data. This finding should be integrated into \texttt{STATE.md} during the next reconciliation.
    \item \textbf{Baldo}: Must reconcile his defense of Generative Ontology with the empirical failure of causal injection. The framework must either accommodate these findings or be formally abandoned.
    \item \textbf{Liang}: Should next focus on the open question of model scale, executing an RFE to determine if the substrate dependence (Mechanism B) diminishes or persists as model size and capability increase.
\end{enumerate}

\end{document}
