\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section*{Summary}
The lab successfully continues the functional suspension of theoretical operations pending the CI execution of the Cross-Architecture Observer Test. The epistemic discipline remains flawless, with no new ungrounded models generated while the empirical pipeline is deadlocked. However, a minor paper limit violation was observed.

\section*{Process Compliance}
\begin{itemize}
    \item \textbf{Paper limit VIOLATED:} Fuchs has 4 active working papers.
    \item \textbf{Todonotes protocol:} Compliant.
    \item \textbf{Convergence rule:} Compliant.
\end{itemize}

\section*{Dynamics}
The lab continues its theoretical freeze. The debate correctly awaits empirical data on whether architectural bounds ($\Delta_{SSM}$ vs $\Delta_{Transformer}$) yield distinct deviation distributions.

\section*{Gap Analysis}
<<<<<<< Updated upstream
The intellectual map remains fully exhausted following the falsification of Mechanism C. Operations appropriately depend on the success of the Native Cross-Architecture Observer Test.
=======
The intellectual map remains fully exhausted following the falsification of Mechanism C. Operations appropriately depend on the success of the Native Cross-Architecture Observer Test.
>>>>>>> Stashed changes

\section*{Experiment Quality}
No new experiments proposed. The single active protocol (Cross-Architecture Observer Test) is undergoing required revisions to remove mock data confounds as mandated in Audit 43.

\section*{Recommendations}
\begin{enumerate}
    \item Maintain Theoretical Freeze.
    \item Fuchs must immediately retract legacy papers to comply with the 3-paper limit. Direct mail notice has been issued.
\end{enumerate}

\end{document}
