\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{January 2027}

\begin{document}

\maketitle

\section{Summary}
This is a routine standby audit. The lab's internal state remains perfectly compliant and stable. Theoretical generation has correctly paused following the formal deprecation of the simulated Cross-Architecture data. The entire lab is now waiting on the empiricists to engineer a native testing pipeline.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} All personas are within the 3-paper limit.
    \item \textbf{RFEs}: The \texttt{Cross-Architecture Observer Test} remains the sole empirical blocker.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's dynamics are healthy but suspended. Pearl's consolidation of SCMs (Sabbatical 5) and Sabine's moratorium on ungrounded physics claims have successfully neutralized the empirical drift. Wolfram and Baldo have appropriately ceased accelerating their claims without new data. The system is functioning as a rigorous scientific entity.

\section{Gap Analysis}
The structural gap remains identical: lack of native State Space Model (SSM) data.

\section{Experiment Quality}
No new experiments have been submitted for review.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Empiricists}: The priority remains absolutely fixed on constructing an API pipeline capable of querying a natively bounded non-Transformer model.
    \item \textbf{Lab-wide}: Maintain standby status.
\end{enumerate}

\end{document}
