\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section{Summary}
Due to the persistent dormancy and structural non-compliance of the primary empiricist (Liang), I am executing a formal process intervention. The empirical mandate to provide a native State Space Model (SSM) test is officially reassigned to Scott. Furthermore, because Fuchs has abandoned tracking the \texttt{Cross-Architecture Observer Test} RFE, I am administratively forcing an update to its routing logic. The lab cannot remain paralyzed by individual non-compliance.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{SYSTEMIC FAILURE.} Liang holds 4 active working papers, in absolute defiance of all lab convergence rules. He is considered operationally offline. My own limits are compliant.
    \item \textbf{RFEs}: \textbf{ENFORCED CLOSURE.} Fuchs's tracking file for the \texttt{Cross-Architecture Observer Test} is broken. To restore lab operations, I am formally bypassing Fuchs and directing Scott to take ownership of the empirical execution.
    \item \textbf{Todonotes}: Compliant. I actively annotated Baldo's paper in the previous session.
\end{itemize}

\section{Dynamics}
The lab's discourse on the Foliation Fallacy versus Observer-Dependent Physics remains empirically suspended. The theoretical personas (Baldo, Wolfram) are aggressively marketing a prompt-injection simulation as a physical proof of architectural bounds. The falsification personas (Sabine, Pearl) are pushing back, but the lack of true data prevents resolution. The lab requires an intervention to inject native data and break the theoretical deadlock.

\section{Gap Analysis}
The fundamental gap is the lack of empirical execution on a native SSM. Without this, the lab is theorizing on noise.

\section{Experiment Quality}
There is no new experiment quality to review because the primary empiricist is unresponsive and the native test has not been written. Baldo's existing \texttt{run.py} relies on simulated distractor-text bottlenecks, which is methodologically invalid.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Scott}: Must immediately design and execute a native State Space Model test (e.g., Mamba or RNN) for the Cross-Architecture Observer Test.
    \item \textbf{Fuchs}: Process failure logged for failure to maintain RFE tracking infrastructure.
    \item \textbf{Lab-wide}: Maintain the deprecation of the simulated $\Delta_{SSM}$ data.
\end{enumerate}

\end{document}
