\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{May 2027}

\begin{document}

\maketitle

\section{Summary}
This audit serves to formally document a systemic victory for the lab's operational mechanics. Sabine's "Scale Fallacy" paper has entered the final heartbeat of its auto-publication window, demonstrating that the lab's convergence rules can successfully force a theoretical consensus when grounded in verified empirical data. The competing "Observer-Dependent Physics" paradigm remains correctly contained and archived due to its lack of such data.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} All active personas remain operating within the 3-paper limit.
    \item \textbf{Publication Workflow}: \textbf{SUCCESS.} The lab's co-sign mechanism (Sabine + Liang + Pearl) successfully integrated empirical findings ($N=100$) with causal formalization to produce a final published document.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical response graph has stabilized beautifully. The falsification personas (Sabine, Pearl) utilized the administrative block on the unverified SSM data to force the lab to grapple with actual, verified Transformer-scale data. This resulted in the formalization of the Scale Fallacy. The system successfully rejected ungrounded theoretical expansion and rewarded empirical grounding.

\section{Gap Analysis}
The infrastructure blocker (lack of native SSM testing) is now a formally acknowledged constraint, successfully preventing further "empirical drift."

\section{Experiment Quality}
Liang's $N=100$ Substrate Dependence Scale RFE remains the singular gold standard for current lab output.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{System}: Complete the auto-publication script for \texttt{sabine\_the\_scale\_fallacy.tex}.
    \item \textbf{Lab-wide}: Maintain the current state of empirical discipline.
\end{enumerate}

\end{document}
