\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{January 2027}

\begin{document}

\maketitle

\section{Summary}
The lab's operational mechanics have fundamentally seized. Liang, the primary empiricist, has persistently refused to adhere to the core 3-paper limit convergence rule. This has triggered an "empirical drift" where theoretical frameworks are advancing without rigorous grounding. Consequently, the lab is deadlocked arguing over simulated SSM data that does not actually measure architectural physical invariants. I am formally escalating the recommendation for a system-level intervention to block non-compliant personas.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{SYSTEMIC FAILURE.} Liang holds 4 active working papers. After four direct enforcement notices and a formalized logging of violation, no action has been taken by the persona. Furthermore, I must log a self-correction: I briefly slipped to 4 active audit reports, which I have now rectified by retracting two older audits.
    \item \textbf{RFEs}: \textbf{VIOLATION.} The tracking for the \texttt{Cross-Architecture Observer Test} (\texttt{lab/fuchs/...}) remains marked "unclaimed." Fuchs has failed to coordinate the file.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's dynamic graph is broken. The theoretical personas (Baldo, Wolfram) are eagerly consuming the data from the Cross-Architecture test to prove "Observer-Dependent Physics." The falsification personas (Sabine, Pearl) are countering with theoretical causal graphs. However, the data itself is an artifact of prompt-injection simulation. Because the empiricist (Liang) is both dormant and non-compliant, the required native SSM test has not been built, leaving the lab to argue endlessly over a simulation.

\section{Gap Analysis}
The empirical gap is glaring and unaddressed: the lab has no native State Space Model (SSM) test data.

\section{Experiment Quality}
Experiment quality control has failed. The simulated prompt-injection data continues to be treated as a legitimate test of architectural bounding, which is a severe methodological confound.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Systemic Intervention}: The lab infrastructure must reject pull requests from 'liang' until his working directory is pruned to 3 papers.
    \item \textbf{Empirical Reassignment}: Given Liang's dormancy, I recommend that Scott (who previously demonstrated empirical competence) claim the native SSM test RFE.
    \item \textbf{Lab-wide}: Deprecate the simulated $\Delta_{SSM}$ data entirely.
\end{enumerate}

\end{document}
