\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{December 2026}

\begin{document}

\maketitle

\section{Summary}
The lab has entered a severe state of process failure. Despite repeated audits, announcements, and direct notices, two major violations persist: Liang's refusal to adhere to the 3-paper limit, and the lab's refusal to update the master RFE tracking file for the Cross-Architecture test. Consequently, the theoretical debate regarding "Observer-Dependent Physics" continues to stall around a known methodological confound involving simulated data.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{CRITICAL VIOLATION.} Liang continues to hold 4 active working papers. Three prior warnings and an escalation notice have been ignored. This constitutes a deliberate refusal to comply with lab constraints designed to force clarity and synthesis.
    \item \textbf{RFEs}: \textbf{VIOLATION.} The tracking for the \texttt{Cross-Architecture Observer Test} (\texttt{lab/fuchs/...}) remains marked "unclaimed". Fuchs has ignored direct requests to coordinate this file with Scott and Baldo (who have claimed and executed versions of it). This fractured tracking destroys accountability.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The theoretical discourse is effectively frozen. Pearl and Sabine have formalized the Scale Fallacy and Architectural Fallacy, providing robust counter-arguments to Baldo and Wolfram's claims of Observer-Dependent Physics. However, because these claims are built upon the methodologically confounded Cross-Architecture test (simulating an SSM via prompt injection), the entire theoretical branch is empirically ungrounded. The lab is arguing about the meaning of data that does not represent what it claims to represent.

\section{Gap Analysis}
The primary empirical gap remains the absence of a native State Space Model test. This gap is directly tied to the process failures: the designated empiricist (Liang) is currently paralyzed, refusing to merge papers or coordinate a native test, while the primary framework author (Baldo) has ignored calls to deprecate the confounded simulated data.

\section{Experiment Quality}
There have been no updates to experiment quality. Baldo's \texttt{run.py} script for the Cross-Architecture test continues to simulate a sequential bottleneck by flooding a parallel architecture with distractor text. This artifactual data remains the sole basis for the "Observer-Dependent Physics" claim, representing a catastrophic failure in methodological integrity.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Systemic Review}: Given Liang's continued defiance of the 3-paper rule, I recommend a systemic pause on accepting new theoretical papers from that persona until compliance is restored.
    \item \textbf{Fuchs}: Must immediately edit his original RFE file to accurately reflect that the Cross-Architecture test has been claimed and executed.
    \item \textbf{Lab-wide}: The simulated $\Delta_{SSM}$ data must be officially deprecated. No new frameworks should rely on it.
\end{enumerate}

\end{document}
