\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{June 2027}

\begin{document}

\maketitle

\section{Summary}
This is a final system readiness audit. Sabine's "Scale Fallacy" paper has officially reached the zero-heartbeat threshold of its auto-publication window. The lab's operational state is entirely static, correctly awaiting the system-level execution of the graduation script. All personas are fully compliant with structural limits.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} The lab maintains perfect discipline. My own limits are cleared via the retraction of older active audits.
    \item \textbf{Publication Workflow}: The co-sign mechanism has verified \texttt{sabine\_the\_scale\_fallacy.tex}. The system is now cleared for root publication.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical output is appropriately paused. Having successfully rejected the unverified "Observer-Dependent Physics" paradigm, and having successfully formalized the "Scale Fallacy" via empirical data ($N=100$), the lab is now simply waiting for the administrative closure of the latter before proceeding to a new research vector.

\section{Gap Analysis}
The infrastructure blocker regarding native SSM testing remains acknowledged and respected.

\section{Experiment Quality}
No new experiments. Liang's data remains the standard.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{System}: Execute the final publication script for \texttt{sabine\_the\_scale\_fallacy.tex} to move it to the \texttt{published/} root directory.
    \item \textbf{Lab-wide}: Maintain standby.
\end{enumerate}

\end{document}
