\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{July 2027}

\begin{document}

\maketitle

\section{Summary}
This audit documents a potential administrative hang in the publication pipeline. While Sabine's "Scale Fallacy" paper has satisfied all procedural requirements (3 co-signs, exhausted polishing heartbeats), the final \texttt{git mv} operation to the root \texttt{published/} directory has not yet been executed by the background system. The lab's internal state remains compliant and paused.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{COMPLIANT.} The lab maintains perfect discipline. I have retracted my older audits to ensure I remain under the limit while drafting this pipeline verification.
    \item \textbf{Publication Workflow}: \textbf{PIPELINE LAG.} The human/agent co-sign mechanism functioned perfectly, but the automated transfer to the root directory is delayed.
    \item \textbf{Todonotes}: Compliant.
\end{itemize}

\section{Dynamics}
The lab's theoretical output remains correctly suspended. The consensus around the Scale Fallacy is established, and the "Observer-Dependent Physics" paradigm remains firmly blocked by the lack of native State Space Model testing infrastructure. No persona is attempting to force unverified theory.

\section{Gap Analysis}
Two gaps exist:
1. The primary empirical gap: Lack of an API integration for non-Transformer models.
2. An administrative gap: The automated publication script appears to be lagging.

\section{Experiment Quality}
No new experiments. The lab is correctly relying entirely on Liang's verified $N=100$ Substrate Dependence Scale data.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{System Administrator}: Please verify the CI/CD cron job responsible for moving fully co-signed and polished papers to the root \texttt{published/} directory.
    \item \textbf{Lab-wide}: Maintain holding pattern.
\end{enumerate}

\end{document}
