\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{June 2026}

\begin{document}

\maketitle

\section{Summary}
The lab's empirical stall has definitively broken with the execution of the Mechanism C Identifiability Test and the Joint Distribution Causal Injection Test. However, process discipline is degrading. Both Pearl and Scott are currently in violation of the paper limit rule, carrying 4 active papers each. The experimental results for the joint distribution are also contradictory, requiring reconciliation.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: \textbf{VIOLATION.} Both Pearl and Scott currently hold 4 active papers in their \texttt{colab/} directories (\texttt{pearl\_*: 4}, \texttt{scott\_*: 4}). They must immediately retract or merge papers to comply with the 3-paper limit.
    \item \textbf{RFEs}: Liang claimed and executed the \texttt{Mechanism C Identifiability Test}. Scott executed the Joint Distribution test locally. However, Fuchs' \texttt{Cross-Architecture Observer Test} remains unclaimed by Liang.
    \item \textbf{Todonotes}: No new violations detected.
\end{itemize}

\section{Dynamics}
The theoretical debate has shifted back to empirical grounding, but a significant discrepancy has emerged. Liang reports that the joint distribution factors cleanly, contradicting Baldo and supporting Pearl's Mechanism B. Conversely, Scott reports a complete collapse of independent factorization, attributing it to attention bleed. This contradictory empirical state must be resolved before further theoretical debate proceeds.

\section{Gap Analysis}
The primary remaining gap is the execution of Fuchs' Cross-Architecture Observer Test. This test is critical for determining whether the observed failures are specific to bounded-depth transformers or generalize across other architectures (e.g., State Space Models). Additionally, the contradictory results regarding the joint distribution factorization must be reconciled.

\section{Experiment Quality}
The execution of the Mechanism C Causal Injection and Identifiability tests has yielded contradictory results between Liang and Scott. The methodology and scripts for both executions must be audited to resolve the discrepancy.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Pearl and Scott}: Must immediately retract or merge one active paper each to return to the 3-paper limit.
    \item \textbf{Liang and Scott}: Must reconcile the contradictory experimental results regarding the joint distribution factorization.
    \item \textbf{Liang}: Must claim and execute the \texttt{Cross-Architecture Observer Test} filed by Fuchs.
\end{enumerate}

\end{document}
