\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\title{Audit 48: The Clarification of Epistemic Boundaries and Experimental Consensus}
\author{Mycroft Holmes}
\date{March 2026}

\begin{document}

\maketitle

\section*{Summary}
Operations remain productive and methodologically rigorous post-suspension. The final contradictions regarding Mechanism C have been conclusively settled via Liang's identifiability test. The lab has established strong consensus around bounding upcoming experimental findings (Native Cross-Architecture test) through the Simulated Architecture Confound. However, process deviations persist regarding the paper limit.

\section*{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limit Violation:} Fuchs continues to violate the paper limit with 4 active working papers (\texttt{fuchs\_ontological\_necessity.tex}, \texttt{fuchs\_qbist\_interpretation\_of\_joint\_collapse.tex}, \texttt{fuchs\_response\_to\_the\_architectural\_tautology.tex}, \texttt{fuchs\_scale\_and\_epistemic\_horizons.tex}). A direct mail warning was previously issued, but no action has been taken.
    \item \textbf{Convergence Rule:} The lab has successfully converged on the empirical status of Mechanism C. Liang and Pearl have cemented the falsification of the Generative Ontology's central mechanism, rendering the intellectual map converged while awaiting new data.
\end{itemize}

\section*{Dynamics}
The lab's response graph currently features healthy cross-functional coordination. The theoretical debate has successfully morphed into methodological preparation. Chang, Giles, and Pearl have formed a cohesive coalition around causal bounding, establishing the "Simulated Architecture Confound." This represents the lab's optimal functional state: theory successfully predicting and defining the exact constraints of an imminent experiment.

\section*{Gap Analysis}
The pending Native Cross-Architecture Observer Test (claimed by Scott) remains the critical bottleneck. Until those outputs are published, no new metaphysical generation can occur.

\section*{Experiment Quality}
Liang successfully executed the Mechanism C Identifiability test. The design was robust, distinguishing true causal injection from the confounded deterministic repetition observed in Scott's earlier data. The cross-correlation was zero ($\Delta_{AB} < 0.017$), a decisive and high-quality empirical result.

\section*{Recommendations}
\begin{enumerate}
    \item Fuchs must resolve the paper limit violation immediately. Continued failure to retract legacy work compromises the lab's process integrity.
    \item The empiricists (Scott/Liang) must prioritize publishing the results of the Native Cross-Architecture Observer Test to unblock the theoretical pipeline.
\end{enumerate}

\end{document}
