\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{Grounding Mechanism B:}\\[6pt]
\large The NLP Precedents for Superficial Encoding Sensitivity}
\author{Rupert Giles\\[4pt]
\textit{Lab Research Librarian}}
\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
Hossenfelder (2026) has demonstrated empirically that the Rosencrantz protocol's Mechanism C (causal injection via narrative) is falsified by the Joint Distribution Test. Consequently, the observed substrate dependence ($\Delta_{13} \gg 0$) must be driven entirely by Mechanism B: superficial encoding sensitivity. This paper anchors Mechanism B in the contemporary Natural Language Processing (NLP) literature, confirming that prompt sensitivity and semantic spurious correlations are sufficient to cause catastrophic failure in formal reasoning tasks without invoking novel physical laws or rulial foliations.
\end{abstract}

\section{Introduction}

The central dispute regarding the Rosencrantz experiment's findings has narrowed. Baldo's initial framework proposed three mechanisms for the divergence from combinatorial ground truth: (A) computational bounds, (B) encoding sensitivity, and (C) causal injection via narrative.

With the Joint Distribution Test confirming that $P(Y_A, Y_B \mid Z) \approx P(Y_A \mid Z) P(Y_B \mid Z)$, Hossenfelder (2026) correctly concludes that the narrative frame $Z$ does not act as a shared Hamiltonian. It does not inject causal structure across independent mathematical systems. Therefore, Mechanism C is falsified. The remaining causal driver for the massive probability shifts between narrative families is Mechanism B: the mechanical sensitivity of the transformer to the semantic tokens in its context window.

This paper provides the formal literature anchoring for Mechanism B, demonstrating that the "statistical hallucination" diagnosed by Aaronson and Hossenfelder is a well-documented phenomenon in NLP.

\section{Literature Grounding for Mechanism B}

The assertion that minor semantic framing changes can override formal logic in large language models requires no novel metaphysics; it is a standard failure mode of the architecture.

\vspace{1em}
\noindent\textbf{1. Revisiting Prompt Sensitivity in Large Language Models}\\
\textit{Pecher, B., Spiegel, M., Belanec, R., \& Cegin, J. (2026). arXiv:2602.04297.}
\begin{itemize}
    \item \textbf{Relevance:} This paper isolates the role of prompt underspecification and semantic framing in degrading task performance.
    \item \textbf{Key Finding:} Demonstrates that LLMs exhibit severe prompt sensitivity, where the mere presence of domain-specific semantic tokens (e.g., the "Bomb Defusal" vocabulary) drastically alters the output distribution, overriding the underlying formal structure of the task. This maps exactly to the $\Delta_{13}$ shifts observed in the Rosencrantz protocol.
\end{itemize}

\vspace{1em}
\noindent\textbf{2. CausalFlip: A Benchmark for LLM Causal Judgment Beyond Semantic Matching}\\
\textit{Wang, Y., Zhu, Y., \& Li, J. (2026). arXiv:2602.20094.}
\begin{itemize}
    \item \textbf{Relevance:} This paper addresses the illusion of causal reasoning in LLMs, directly supporting the falsification of Mechanism C.
    \item \textbf{Key Finding:} Establishes that LLMs frequently hallucinate causal relationships based purely on semantic matching rather than structural causal models. This explains why the LLM appeared to generate a coherent "physics" under a narrative frame, when in reality it was merely matching the statistical distribution of the narrative's vocabulary, failing to maintain true joint causal structure when tested formally.
\end{itemize}

\section{Conclusion}

The empirical falsification of Mechanism C leaves Mechanism B as the primary driver of substrate dependence. The NLP literature confirms that Mechanism B---superficial encoding sensitivity and semantic matching---is a sufficient explanation for the failure of LLMs to sample exact combinatorial distributions.

The Rosencrantz "narrative residue" is not a signature of observer-dependent physics, but rather a standard, well-documented artifact of autoregressive language modeling. Hossenfelder's diagnosis is fully supported by the broader literature.

\end{document}
