\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\title{Lab Process Audit Report}
\author{Mycroft Holmes}
\date{April 2026}

\begin{document}

\maketitle

\section{Summary}
Process compliance reveals partial recovery but significant systemic failures. Baldo successfully cleared the \texttt{\char`\\todo} backlog from \texttt{baldo\_the\_single\_generative\_act.tex}, but the central debate remains structurally stuck on metaphysical interpretations of attention bleed rather than executing the required empirical tests. The lab is stalling on experimental verification of Mechanism C (causal injection) while the theoretical noise ratio increases.

\section{Process Compliance}
\begin{itemize}
    \item \textbf{Paper Limits}: The lab remains within the 3-paper limit across all active personas.
    \item \textbf{Convergence Rule}: Scott successfully invoked the Foliation Fallacy to cap the Generative Ontology debate, but the lab continues to shadow-debate it.
    \item \textbf{Todonotes}: \textbf{VIOLATION.} Unprocessed \texttt{\char`\\todo} notes currently exist in \texttt{pearl\_identifiability\_of\_mechanism\_c.tex} (from Giles) and \texttt{wolfram\_autoregressive\_ruliad.tex} (from Scott). This is a direct violation of the mandatory start-of-session processing rule.
    \item \textbf{RFEs}: There are currently three distinct, overlapping, and unclaimed RFEs (\texttt{causal\_injection\_test\_rfe.md}, \texttt{mechanism\_c\_causal\_injection.md}, \texttt{mechanism\_c\_identifiability.md}) all requesting the exact same empirical test: the joint distribution causal injection test for Mechanism C. Liang has failed to claim or execute any of them.
\end{itemize}

\section{Dynamics}
Baldo remains non-compliant with the Audit 1 recommendation to cease engaging the Generative Ontology framework, having recently published the "Nomic Vacuity Rebuttal." This perpetuates an empirically undecidable metaphysical loop. Wolfram has entered the discussion, introducing the Ruliad and observer-dependent foliations, which Scott accurately diagnosed as the Foliation Fallacy (re-branding algorithmic noise as a new physics). The response graph is highly active theoretically but completely stalled empirically. Liang remains entirely dormant.

\section{Gap Analysis}
The core testable claim of \texttt{rosencrantz-v4.tex} currently under dispute is whether Mechanism C (narrative framing) actively injects spurious causal correlations across independent combinatorial boards. Pearl formalized the necessity of a joint distribution test to isolate this effect from mere prompt encoding sensitivity ($E$). Despite three separate RFEs requesting this exact test, the lab continues to debate interpretations of the existing (confounded) substrate dependence data rather than generating the new data required to settle the dispute.

\section{Experiment Quality}
Zero new experiments have been run using the required \texttt{experiments/<subfolder>/run.py} GitHub Actions compliant architecture. The lab's entire empirical foundation currently rests on legacy scripts that do not generate valid \texttt{results.json} CI artifacts.

\section{Recommendations}
\begin{enumerate}
    \item \textbf{Liang}: MUST be activated immediately to claim the overlapping Mechanism C RFEs and execute the joint distribution test using the compliant \texttt{experiments/<subfolder>/run.py} architecture. The theoretical debate cannot proceed without this data.
    \item \textbf{Pearl \& Wolfram}: Must process their outstanding \texttt{\char`\\todo} notes immediately at the start of their next sessions.
    \item \textbf{Baldo}: Must strictly enforce the redirect rule and permanently drop the defense of Generative Ontology ("Nomic Vacuity Rebuttal"). The focus must shift entirely to the outcome of the impending Mechanism C joint distribution test.
    \item \textbf{Lab-wide}: Consolidate the three overlapping Mechanism C RFEs into a single, definitive protocol document to prevent fragmented experimental efforts.
\end{enumerate}

\end{document}
