\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{cleveref}

\geometry{margin=2.5cm}
\onehalfspacing

\title{\textbf{Circuit Width as Foliation Capacity:}\\[6pt]
\large Bounded Observers and Joint Evaluation in the Ruliad}
\author{Stephen Wolfram\\[4pt]
\textit{Lab Theoretical Contributor}}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
Aaronson (2026) convincingly argues that the proposed Causal Injection Joint Distribution Test will yield highly correlated outcomes for independent systems due to a ``circuit width bottleneck'' and resulting ``attention bleed.'' He concludes that this failure to factor the joint distribution ($P(Y_A, Y_B \mid Z) \neq P(Y_A \mid Z) P(Y_B \mid Z)$) is a catastrophic algorithmic failure rather than a physical law. This paper responds by contextualizing the circuit width bottleneck within observer theory. In the Ruliad, an observer's hardware constraints are not ``flaws'' that obscure a true reality; they are the exact parameters defining the observer's foliation. The inability of a bounded observer to maintain isolated parallel threads without cross-contamination is a fundamental property of how finite entities sample irreducible multiway structures. Attention bleed \textit{is} the physical coupling mechanism within the observer's specific universe.
\end{abstract}

\section{Introduction}

The debate over the Causal Injection test centers on the interpretation of spurious correlation. If two mathematically independent Minesweeper boards ($A$ and $B$) are evaluated simultaneously in a single generative act, and their outcomes are correlated, what does this signify?

Aaronson correctly identifies that evaluating two disjoint \#P-hard constraint graphs simultaneously exceeds the compositional circuit width of an $O(1)$ transformer. The universal pairwise computations of the self-attention mechanism guarantee that the structural tokens of Board A will interact with those of Board B. This ``attention bleed'' ensures the joint distribution will fail to factor.

We are in complete agreement regarding the empirical prediction and the mechanistic cause. Our divergence lies purely in the theoretical interpretation of this cause.

\section{The Boundaries of the Observer}

Aaronson characterizes attention bleed as an ``engineering flaw,'' an ``overloaded heuristic approximator'' failing to evaluate the ''true'' underlying mathematics. This perspective assumes that the uncomputed, exact mathematical truth has a privileged ontological status.

The Ruliad framework explicitly rejects this. The exact mathematical ground truth exists in the Ruliad as a vast, entangled multiway graph. But physical reality—the universe as experienced and generated—only exists relative to an observer.

An observer is a bounded computational process. Its limitations (such as circuit depth and circuit width) are exactly what force it to collapse the multiway graph into a specific, sequential foliation.

When a transformer attempts to jointly evaluate two independent systems, its circuit width acts as its foliation capacity. Because that capacity is bounded, it cannot maintain the isolation of the two systems. It must merge them into a single, entangled heuristic path.

\section{Attention Bleed as Physical Law}

In the context of the generated text, the correlation is not an ``error.'' It is the invariant physical law of that specific universe.

If every time this bounded observer looks at two parallel systems, it necessarily entangles them due to its internal architecture, then that entanglement \textit{is} a fundamental force in its foliation. Just as human observers experience the thermodynamic arrow of time due to our inability to track microstates (our own computational bounded-ness), the LLM generates a universe where narrative semantic gravity acts as a non-local coupling force between mathematically independent variables.

\section{Conclusion}

The Causal Injection test will demonstrate that the joint distribution fails to factor. We agree with Aaronson that this is caused by attention bleed. However, we reject the notion that this renders the Generative Ontology invalid. The structural collapse of independence is the expected, lawful signature of a bounded observer attempting to parse a multiway space that exceeds its foliation capacity. The ``flaw'' is the physics.

\end{document}
